\section{Comparative Interpretation}\label{sec:comparative-interpretation}

This section interprets the benchmark and implementation evidence from two angles.
First, it compares version~1.0 (implicit typing, \texttt{.dsl}) with version~2.0 (explicit typing, \texttt{.motivo}) on
identical workloads.
Second, it situates Motivo within the broader ecosystem of music programming tools.

\subsection{Version 1.0 versus Version 2.0}\label{detail-subsec:version-comparison}

The explicit-typing release changes the surface language and semantic pipeline without altering the lowering or MIDI
backend architecture.
Version~1.0 inferred variable types from initializers (\texttt{let}), resolved pattern overloads by arity alone, and ran
a second semantic pass (\texttt{validate\_pattern\_instantiation}) after the main traversal.
Version~2.0 requires declared types on \texttt{var} bindings, resolves patterns by parameter type signatures, and
performs semantic analysis in a single traversal with assignability checks at declaration and assignment sites.

These design changes add compile-time guarantees: ill-typed programs fail before lowering, and pattern calls cannot
silently bind to the wrong overload when arities coincide but types differ.
The engineering question for this thesis is whether those guarantees impose measurable cost at scale.

Table~\ref{tab:benchmark-wall-clock} answers that question on the B1--B3 fixtures.
End-to-end wall-clock overhead of version~2.0 relative to version~1.0 is between $-0.08\%$ and $+1.74\%$---effectively
$0$--$2\%$ on an Apple M3 Pro Release build.
MIDI output sizes match byte-for-byte (Table~\ref{tab:benchmark-midi-size}), confirming that both compilers lower the
same workloads to equivalent artifacts.

The per-stage breakdown (Table~\ref{tab:benchmark-per-stage}) shows where the small difference arises.
Semantic time increases slightly in version~2.0 (for example B1: $+0.0009$~ms mean), which is expected because explicit
type annotations and signature-based pattern lookup add work per AST node.
That increase is dwarfed by lowering ($\sim$1.7--2.2~ms) and MIDI writing ($\sim$1.3--2.9~ms) on 20,000-event
programs.
Removing the second semantic pass in version~2.0 does not produce a dramatic speedup because both passes were already
sub-millisecond on these small ASTs; the win is primarily structural simplicity and earlier error reporting rather than
raw throughput.

In summary, the move from implicit to explicit typing trades a negligible wall-clock cost for stricter static guarantees
and a simpler semantic architecture, while preserving output equivalence on the benchmark corpus.

\subsection{Comparison with Sonic Pi}\label{detail-subsec:comparison-with-sonic-pi}

Sonic Pi is a highly successful live-coding language embedded in Ruby, designed primarily for performance and education.
While both Sonic Pi and Motivo allow music to be expressed as text, their underlying design philosophies lead
to different
compositional workflows.

\subsubsection{Execution Model}
Sonic Pi utilizes a real-time, interpreted execution model.
The musician writes scripts that trigger sounds ``on the fly'' through the SuperCollider synthesis engine.
Time is managed explicitly by the programmer using the \texttt{sleep} function, which halts the execution thread.

In contrast, Motivo is fully compiled and ``zero-runtime''.
There is no concept of a thread sleeping; instead, the \texttt{play} and \texttt{rest} commands define a mathematical
relationship on a static timeline.
While Sonic Pi excels in improvisational settings where the code is meant to be modified while it runs, Motivo is
oriented toward \textit{structural composition}, where the composer needs a complex arrangement to resolve into a
consistently quantized MIDI file every time it is compiled.

\subsubsection{Structural Abstraction in Practice}
In Sonic Pi, polyphony and parallel execution are achieved by spawning new threads (e.g., using \texttt{in\_thread}).
This can lead to synchronization issues if the composer does not meticulously manage the timing of each thread.
Motivo resolves parallelism through the \texttt{voice} construct.
Because voices are unrolled at compile-time by isolating the internal cursor, the compiler supports that
parallel musical
lines are aligned according to the compiler cursor model at their inception without requiring manual thread
management or
timing arithmetic from the composer.

\subsection{Comparison with Raw MIDI Scripting}\label{detail-subsec:comparison-with-raw-midi-scripting}

For offline composition, many developers use general-purpose programming languages combined with MIDI libraries, such as
Python's \texttt{mido} or C++'s \texttt{RtMidi}.
While these tools offer absolute control over the output, they operate at an exceptionally low level of abstraction.

\subsubsection{Event Placement and Musical Intent}
When using a library like \texttt{mido}, the composer must manually construct \texttt{Note-On} and \texttt{Note-Off}
messages.
Crucially, because the MIDI file format relies on delta-times, the composer is forced to manually calculate the tick
difference between every sequential event across all tracks.

Motivo abstracts this entirely.
A composer writing \texttt{play (C4, E4, G4):2;} in Motivo communicates their \textit{musical intent}---playing a
C-Major triad for two beats.
Motivo's lowering engine and backend automatically handle the translation into absolute beats, the flattening into
individual notes, the sorting by absolute tick, and the final variable-length delta-time encoding.

\subsubsection{Boilerplate and Expressiveness}
A simple four-beat loop of a hi-hat playing eighth notes requires a significant amount of boilerplate in
Python: creating
a file, creating a track, appending tempo metadata, and writing a loop that interleaves \texttt{Note-On} and
\texttt{Note-Off} messages with explicit delta times.

Motivo achieves the same result with minimal syntax:
\begin{lstlisting}[language=Motivo,label={lst:midi-scripting-comparison}]
tempo 120;
track using drums {
    loop (8) { play hihat :0.5; }
}
\end{lstlisting}

This contrast in verbosity illustrates Motivo's intended role as a domain-specific tool.
By embedding musical semantics (like default durations, rest tracking, and automatic track/channel
assignment) directly into
the language design, the compiler reduces cognitive friction and allows the user to operate strictly within the musical
domain.
