\section{Compilation Time}\label{sec:compile-time}

Table~\ref{tab:benchmark-wall-clock} reports end-to-end wall-clock means for the three benchmark programs on version~1.0
and version~2.0.
All pairs produced byte-identical MIDI output (Table~\ref{tab:benchmark-midi-size}), so the timing difference isolates
compiler implementation changes rather than workload variation.
Version~2.0 adds between $-0.01$~ms and $+0.14$~ms relative to version~1.0, or roughly $0$--$2\%$ overhead on these
workloads.

\begin{table}[htbp]
  \centering
  \footnotesize
  \renewcommand{\arraystretch}{1.25}
  \begin{tabular}{lrrrrr}
    \toprule
    \textbf{Program} & \textbf{LOC} & \textbf{Note events} & \textbf{v1 mean ms} & \textbf{v2 mean ms} &
    \textbf{v2/v1} \\
    \midrule
    B1 (\texttt{single\_loop\_20000}) & 8 & 20,000 & $7.92 \pm 0.41$ & $8.06 \pm 0.62$ & 1.017 \\
    B2 (\texttt{chord\_loop\_5000}) & 8 & 9,980 & $5.06 \pm 0.21$ & $5.10 \pm 0.19$ & 1.008 \\
    B3 (\texttt{pattern\_cascade}) & 12 & 20,000 & $8.07 \pm 0.21$ & $8.06 \pm 0.27$ & 0.999 \\
    \bottomrule
  \end{tabular}
  \caption{End-to-end compilation time (version~1.0 \texttt{dslrc} vs.\ version~2.0 \texttt{motivoc}).
    Values are mean~$\pm$~$\sigma$ over 50 measured runs after 5 warmups on an Apple M3 Pro Release build.
  Standard deviations appear in the mean column for compactness.}\label{tab:benchmark-wall-clock}
\end{table}

\begin{table}[htbp]
  \centering
  \footnotesize
  \renewcommand{\arraystretch}{1.25}
  \begin{tabular}{lrrr}
    \toprule
    \textbf{Program} & \textbf{v1 MIDI (B)} & \textbf{v2 MIDI (B)} & \textbf{Match} \\
    \midrule
    B1 & 180,056 & 180,056 & yes \\
    B2 & 84,886 & 84,886 & yes \\
    B3 & 180,056 & 180,056 & yes \\
    \bottomrule
  \end{tabular}
  \caption{MIDI output size confirms semantic equivalence between compiler versions on each fixture.}%
  \label{tab:benchmark-midi-size}
\end{table}

The largest cases finish in about 8~ms on the stated machine.
Because these times are small, they should be interpreted as local measurements rather than as a portable performance
result.
They are still useful for the thesis because they show that the current native compiler is fast enough to support an
interactive compile loop in this local setup.

Table~\ref{tab:benchmark-per-stage} breaks the in-process work down by pipeline stage.
Parsing and semantic analysis remain below $0.04$~ms regardless of program; lowering and MIDI writing dominate, which
matches the $O(E)$ and $O(E \log E)$ expectations from Section~\ref{detail-subsec:theoretical-complexity}.
The semantic column shows a slight increase in version~2.0 (for example B1: $0.0058$~ms $\rightarrow$ $0.0067$~ms),
consistent with explicit type checking, but the absolute cost is negligible compared with lowering and backend I/O.

\begin{table}[htbp]
  \centering
  \footnotesize
  \renewcommand{\arraystretch}{1.25}
  \begin{tabular}{llrrrr}
    \toprule
    \textbf{Program} & \textbf{Ver.} & \textbf{Parse} & \textbf{Semantic} & \textbf{Lowering} & \textbf{MIDI write} \\
    \midrule
    B1 & v1 & 0.032 & 0.006 & 1.655 & 2.889 \\
    B1 & v2 & 0.034 & 0.007 & 1.665 & 2.867 \\
    \midrule
    B2 & v1 & 0.030 & 0.005 & 0.901 & 1.346 \\
    B2 & v2 & 0.029 & 0.006 & 0.902 & 1.346 \\
    \midrule
    B3 & v1 & 0.032 & 0.006 & 2.194 & 2.759 \\
    B3 & v2 & 0.031 & 0.006 & 2.171 & 2.736 \\
    \bottomrule
  \end{tabular}
  \caption{Mean in-process stage times (ms) with \texttt{MOTIVO\_BENCHMARK=1}.
  Stage sums exclude process spawn and filesystem overhead present in Table~\ref{tab:benchmark-wall-clock}.}%
  \label{tab:benchmark-per-stage}
\end{table}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        motivo/bar,
        width=0.92\textwidth,
        height=6.5cm,
        ymin=0,
        ylabel={Mean compile time (ms)},
        symbolic x coords={B1,B2,B3},
        xtick=data,
        legend style={at={(0.5,-0.16)}, anchor=north, legend columns=2, font=\scriptsize},
        bar width=13pt,
        enlarge x limits=0.15,
      ]
      \addplot[fill=motivoBlue!70, draw=motivoBlue!90] coordinates {
        (B1,7.92)
        (B2,5.06)
        (B3,8.07)
      };
      \addplot[fill=motivoGreen!70, draw=motivoGreen!90] coordinates {
        (B1,8.06)
        (B2,5.10)
        (B3,8.06)
      };
      \legend{Version 1.0, Version 2.0}
    \end{axis}
  \end{tikzpicture}
  \caption{End-to-end wall-clock means for benchmark programs B1--B3 (version~1.0 vs.\ version~2.0).}%
  \label{fig:benchmark-time}
\end{figure}

The near-linear growth in the single-loop case (B1) is consistent with the implementation.
Parsing and semantic analysis see a small source program, while lowering and MIDI writing scale with the number of
generated events.
B3 stresses nested loop structure rather than a flat iteration count; its lowering time exceeds B1 despite
the same event
cap because the lowering engine walks nested control-flow frames.
The backend also sorts events by tick before computing delta-times.
The result is still small at the current safety limit, but the implementation should not be described as unbounded.

\subsection{Execution Speed and Real-Time Viability}\label{detail-subsec:execution-speed-and-real-time-viability}

Because the compiler is implemented as a native C++23 program without reliance on garbage collection or dynamic dispatch
for AST traversal, the constant factors in the time complexity are small in the current measurements.
In practice, the compiler generates compositions containing event streams up to the current implementation limit in a
fraction of a second.

These local measurements are relevant to Motivo Studio because the editor workflow depends on compilation remaining
quick enough for interactive use.
The interactive playground environment (discussed in previous chapters) relies on the compiler's ability to re-compile
the entire source file during interactive compile actions.
Because the execution time for typical musical pieces (ranging from hundreds to thousands of events) sits well below the
latency usually noticed during ordinary editor interaction, the composer experiences interactive auditory and visual
feedback without using an interpreted runtime during playback.
