\section{The Compilation Strategy: Deterministic Unrolling}\label{sec:compilation-strategy}

The proposed DSL is designed as a fully compiled language, distinguishing it from interpreted musical environments.
This design choice is driven by the mandate for a \textit{zero-runtime architecture}, where all musical and logical structures are resolved before any sound is produced or any output file is written.

\subsection{The Zero-Runtime Mandate}

In a zero-runtime system, the complexity of the composition is handled entirely by the compiler.
This means that features such as nested loops, conditional branching, and recursive pattern calls do not exist in the output artifact.
Instead, the compiler performs a process of \textit{deterministic unrolling}.
Every loop is expanded into its constituent iterations, every conditional branch is evaluated and resolved, and every pattern call is inlined.

The result of this process is an Intermediate Representation (IR) that consists of a flat, linear list of atomic musical events.
This architecture provides several benefits:
\begin{itemize}
    \item \textbf{High Performance:} The output artifact is extremely lightweight, as all logical overhead has been removed.
    \item \textbf{Offline Validation:} Semantic errors—such as variable shadowing, type mismatches, or invalid rhythmic math—are caught during the compilation phase, providing immediate feedback to the composer.
    \item \textbf{Portability:} The resulting IR can be easily serialized into multiple formats (such as MIDI or JSON) without requiring a specialized interpreter to read it.
\end{itemize}

\subsection{Temporal Reconstruction of Patterns}

A unique aspect of the compilation strategy is how the temporal span of a pattern is reconstructed.
When a pattern is called, the compiler does not just emit the notes within it; it calculates the total duration occupied by the pattern, including internal gaps and trailing rests.
This is achieved by capturing the state of the internal cursor before and after the pattern's body is unrolled.
The difference between these two points defines the pattern's "footprint" on the timeline, allowing the pattern to be treated as a single, opaque musical unit by the caller.

\subsection{Static Verification of Compositional Logic}

Because the language is compiled, it can enforce constraints that would be difficult to manage in an interpreted system.
The semantic analysis pass verifies that the composition adheres to the physical limits of the target protocol (e.g., ensuring notes are within the MIDI pitch range of 0--127) and that the structural hierarchy is valid (e.g., preventing a \texttt{track} from being defined inside a \texttt{pattern}).
This rigorous validation ensures that the final MIDI artifact is not only syntactically correct but also musically meaningful and interoperable with other professional audio software.
