\section{Deployment and Runtime Configuration}\label{sec:deployment}

Motivo Studio is deployed twice in practice: a \textbf{local Docker Compose stack} that emulates production topology
for development and thesis demonstrations, and a \textbf{production stack} on Railway with Cloudflare DNS and TLS in
front of \texttt{motivo-studio.eu}.
The two environments share the same logical shape---browser to Next.js client, same-origin \texttt{/api/*} proxy to
Express, private database networking, embedded \texttt{motivoc} in the server image---but differ in TLS termination
and gateway components.

\subsection{Local Docker Compose stack}\label{subsec:local-docker}

\texttt{code/docker-compose.yml} defines four cooperating services plus a named Postgres volume:

\begin{itemize}
  \item \textbf{motivo-studio-db} --- PostgreSQL~16 with database \texttt{motivo\_studio}; exposed on host port~5432 for
    optional direct inspection.
  \item \textbf{motivo-studio-server} --- multi-stage image from \texttt{server/Dockerfile} that builds and installs
    \texttt{motivoc} at \texttt{/usr/local/bin/motivoc}, then runs \texttt{node dist/db/migrate.js} followed by
    \texttt{node dist/server.js} on container start (migrations before listen).
  \item \textbf{motivo-studio-client} --- Next.js production build with \texttt{API\_PROXY\_URL} pointing at the server
    service hostname on the Compose network.
  \item \textbf{motivo-studio-gateway} --- nginx container (\texttt{code/nginx/}) listening on ports~80 and~443,
    redirecting HTTP to \texttt{https://motivo-studio.local}, terminating HTTPS with a local self-signed
    certificate, proxying
    \texttt{/api/*} to Express (stripping the prefix) and all other paths to the client (including WebSocket headers for
    hot reload during development builds).
\end{itemize}

Environment variables mirror production semantics: \texttt{PORT}, \texttt{HOSTNAME}, \texttt{COMPILER\_BIN},
\texttt{DATABASE\_URL}, session cookie name/TTL/secret, and \texttt{API\_PROXY\_URL} on the client.
Local Compose sets \texttt{SESSION\_COOKIE\_SECURE=false} because TLS terminates at nginx rather than at Node;
production sets the secure flag on session cookies.

\begin{figure}[htbp]
  \centering
  \motivoscalefig{%
    \begin{tikzpicture}[
        svc/.style={motivo/service, minimum width=3.1cm, minimum height=0.95cm},
        arrow/.style={motivo/arrow, line width=1.3pt}
      ]
      \node[svc, fill=motivoSoft] (browser) {Browser\\\texttt{motivo-studio.local}};
      \node[svc, fill=motivoAmber!18, below=0.75cm of browser] (nginx) {nginx gateway\\HTTPS :443};
      \node[svc, fill=motivoBlue!13, below left=0.85cm and 0.5cm of nginx] (client) {Next.js\\port 3000};
      \node[svc, fill=motivoCyan!16, below right=0.85cm and 0.5cm of nginx] (server) {Express\\port 3001};
      \node[svc, fill=motivoGreen!16, below=0.65cm of server] (compiler) {\texttt{motivoc}};
      \node[svc, fill=motivoSoft, below left=0.85cm and 0.5cm of server] (db) {PostgreSQL};

      \draw[arrow] (browser) -- (nginx);
      \draw[arrow] (nginx) -- node[left, font=\scriptsize\sffamily] {pages} (client);
      \draw[arrow] (nginx) -- node[right, font=\scriptsize\sffamily] {\texttt{/api/*}} (server);
      \draw[arrow] (server) -- (compiler);
      \draw[arrow] (server) -- (db);
    \end{tikzpicture}%
  }
  \caption{Local Docker Compose topology.
  nginx is used only in development; production omits this hop.}%
  \label{fig:deployment-local}
\end{figure}

Docker here is a reproducibility tool: it pins the compiler toolchain, Node runtimes, and service wiring so thesis
results and demos do not depend on ad hoc host installations.

\subsection{Production: Railway, Cloudflare, and domain}\label{subsec:production-railway}

Production (author-confirmed, 2026-06-01) serves \texttt{https://motivo-studio.eu} and
\texttt{https://www.motivo-studio.eu}.

\topic{Railway.} Hosts three services, all in the Amsterdam region:

\begin{itemize}
  \item \textbf{Frontend} --- Docker image from \texttt{code/client/Dockerfile}; public Next.js entrypoint.
  \item \textbf{Backend} --- Docker image from \texttt{code/server/Dockerfile} with embedded
    \texttt{motivoc}; not exposed
    as a separate public host---reachable from the client via internal networking and \texttt{API\_PROXY\_URL}.
  \item \textbf{Database} --- managed PostgreSQL on private Railway networking (not publicly routable).
\end{itemize}

Deploy automation integrates with GitHub: pushes to \texttt{main} run module-scoped CI workflows first; Railway deploys
only after the relevant pipeline succeeds.
Path-scoped watches limit churn---client-only changes redeploy the frontend service; server or compiler changes redeploy
the backend (which rebuilds the embedded compiler).
The server service runs \texttt{node dist/db/migrate.js} as a \textbf{pre-deploy} hook configured in the Railway
dashboard so schema changes apply before new containers accept traffic.

\topic{Cloudflare.} Fronts the zone \texttt{motivo-studio.eu}: apex and \texttt{www} DNS records proxy (orange cloud) to
the Railway Next.js origin with SSL/TLS mode \textbf{Full (Strict)}.
There is no separate API subdomain; browsers use the same origin for pages and \texttt{/api/*}, matching the Next.js
rewrite design locally.

\begin{figure}[htbp]
  \centering
  \motivoscalefig{%
    \begin{tikzpicture}[
        svc/.style={motivo/service, minimum width=3.0cm, minimum height=0.95cm},
        arrow/.style={motivo/arrow}
      ]
      \node[svc, fill=motivoSoft] (browser) {Browser};
      \node[svc, fill=motivoAmber!20, below=0.65cm of browser] (cf) {Cloudflare\\Full Strict TLS};
      \node[svc, fill=motivoBlue!13, below=0.65cm of cf] (client) {Railway client\\Amsterdam};
      \node[svc, fill=motivoCyan!16, below left=0.75cm and 0.3cm of client] (server) {Railway server\\private};
      \node[svc, fill=motivoSoft, below right=0.75cm and 0.3cm of client] (db) {Railway Postgres\\private};

      \draw[arrow] (browser) -- (cf);
      \draw[arrow] (cf) -- (client);
      \draw[arrow] (client) -- node[above, font=\scriptsize\sffamily] {\texttt{/api/*}} (server);
      \draw[arrow] (server) -- (db);
    \end{tikzpicture}%
  }
  \caption{Production request path: Cloudflare terminates public TLS, Railway Next.js proxies API calls to Express on
  private networking, and PostgreSQL is reachable only from the server service.}\label{fig:deployment-production}
\end{figure}

Session secrets, database URLs, and cookie security flags live in Railway environment variables
(\texttt{SESSION\_SECRET}, \texttt{DATABASE\_URL}, \texttt{SESSION\_COOKIE\_SECURE=true}, etc.), analogous to the
Compose \texttt{.env.example} templates in the code repository.

\subsection{CI coupling to deploy}\label{subsec:ci-deploy-gate}

Module CI workflows (compiler, server, client, and this paper) run on GitHub Actions for each push.
Railway's dashboard integration treats a green \texttt{main} pipeline as a deploy gate before promoting new container
images.
This separates quality verification from release: failing tests block production rollout without manual override.
