\section{Source Syntax and Formal Grammar}\label{sec:source-grammar}

Motivo is defined by a formal grammar that ensures syntactic consistency and enables the generation of an
Abstract Syntax Tree (AST).
This grammar is divided into a lexical specification (the scanner) and a context-free grammar (the parser).
The authoritative grammar file in the implementation repository is \texttt{compiler/docs/grammar.bnf}; the
discussion below aligns with that specification at version~2.0.

\subsection{Lexical Grammar}\label{detail-subsec:lexical-grammar}

The lexical layer tokenizes the raw source text into a stream of symbols.
A notable feature of the lexical grammar is the inclusion of musical literals as first-class tokens.
For example, a note literal is defined by the regular expression \verb|[A-G][#b]?[0-8]|, allowing the scanner to
directly recognize pitches such as \texttt{F\#4} or \texttt{Bb2}.
Keywords are reserved for musical structural elements (\texttt{track}, \texttt{pattern}, \texttt{voice}),
temporal modifiers (\texttt{play}, \texttt{from}, \texttt{rest}), control flow (\texttt{for}, \texttt{loop},
\texttt{if}, \texttt{else}), and the declarable type keywords \texttt{int}, \texttt{double}, \texttt{bool},
\texttt{note}, \texttt{seq}, and \texttt{chord}.
The literals \texttt{rest} and drum names (\texttt{kick}, \texttt{snare}, \texttt{hihat}, and related tokens) appear
in expressions and \texttt{play} targets but are not used as variable declaration types.

\subsection{Syntactic Grammar}\label{detail-subsec:syntactic-grammar}

The syntactic grammar follows an EBNF (Extended Backus-Naur Form) structure.
It is designed to enforce the hierarchical relationship between musical components.
The root of a program consists of tempo and time-signature metadata followed by top-level declarations.
Key-signature declarations are not part of the current language and are treated later as future work.

\begin{figure}[htbp]
  \centering
  \begin{lstlisting}[basicstyle=\ttfamily\scriptsize,breaklines=true]
<program>        ::= <header> <top-item>*
<header>         ::= (<tempo-decl> | <signature-decl>)*
<top-item>       ::= <var-decl-stmt> | <pattern-def> | <track-decl>

<track-decl>     ::= "track" IDENT? ("using" <instrument>)? "{" <track-body> "}"
<track-body>     ::= (<stmt> | <pattern-def> | <voice-decl>)*
<voice-decl>     ::= "voice" ("from" <expr>)? "{" <voice-body> "}"

<pattern-def>    ::= "pattern" IDENT "(" <opt-param-list> ")" "{" <stmt-list> "}"
<param>          ::= <type> IDENT
<type>           ::= "int" | "double" | "bool" | "note" | "seq" | "chord"

<var-decl-stmt>  ::= <type> IDENT "=" <expr> ";"
<assign-stmt>    ::= IDENT "=" <expr> ";"
<for-stmt>       ::= "for" "(" <for-init> ";" <for-cond> ";" <for-step> ")" <body-stmt>
<loop-stmt>      ::= "loop" "(" <expr> ")" <body-stmt>
<if-stmt>        ::= "if" "(" <expr> ")" <body-stmt> ("else" <body-stmt>)?

<play-stmt>      ::= "play" <play-target> ";"
<play-target>    ::= <durational-src> <opt-duration> <opt-from>
                 |  <plain-src> <opt-from>
<opt-duration>   ::= (empty | ":" <expr>)
<opt-from>       ::= (empty | "from" <expr>)
  \end{lstlisting}
  \caption{Core Motivo grammar (version~2.0), aligned with \texttt{grammar.bnf}.}\label{fig:grammar-ebnf}
\end{figure}

The grammar allows both braced and unbraced statement bodies in control flow structures, providing a balance between
clarity and brevity.
Crucially, the grammar restricts certain constructs based on their context; for instance, while a \texttt{track} can
contain multiple \texttt{voice} blocks, a \texttt{voice} cannot nest another \texttt{track}, enforcing a one-to-one
mapping between tracks and instrument identities.
The expression sub-grammar supports a full range of arithmetic and logical operators, including a ternary conditional
operator, enabling complex musical logic to be expressed inline.
Pattern parameters require an explicit type before each identifier, and variable declarations at program, track, or
pattern scope use the same \texttt{<type> IDENT = <expr>} form rather than an untyped binding keyword.

\subsection{Evolution from Version 1.0 to Version 2.0}\label{detail-subsec:v1-v2-grammar-evolution}

The Motivo compiler underwent a deliberate language revision between version~1.0 and version~2.0 (mapped in the
repository to git tags \texttt{v1.0.0-base} and \texttt{v2.0.0-explicit-types}).
Version~1.0 followed a more conventional scripting style: bindings used the keyword \texttt{let}, and the semantic
analyzer inferred each variable's type from its initializer.
Pattern parameters were untyped name lists, and the compiler selected among overloaded patterns by \textit{arity}
(parameter count) alone.
A second semantic traversal, \texttt{validate\_pattern\_instantiation}, simulated pattern bodies at call sites to catch
errors that only appeared under specific argument shapes.

Version~2.0 replaces that model with \textit{explicit static typing} at the source level.
The \texttt{let} keyword is removed; declarations are written as \texttt{<type> name = expr}, parsed into
\texttt{VarDeclStatement} nodes with a \texttt{types::Type} field.
Pattern parameters become \texttt{TypedParameter} records, and overload resolution uses the full parameter
\textit{type signature}---not merely arity---via \texttt{find\_visible\_pattern\_by\_signature}.
Semantic analysis is a \textit{single} traversal of the program: pattern symbols are collected when their enclosing
scope is entered, then statements and expressions are checked in one pass; the dedicated instantiation validator is
removed because call-site checking (\texttt{validate\_call}) enforces argument types against the resolved overload.

This evolution trades slightly more verbose source notation for earlier, clearer diagnostics: a type mismatch is
reported at the declaration or call that introduced it, rather than being deferred to a simulated second pass.
The grammar impact is concentrated in declarations and pattern signatures; control flow, \texttt{play}, tracks, and
voices retain the same surface structure, which keeps existing compositional idioms recognizable across versions.
