\section{Conceptual Model: Structural Composition}\label{sec:conceptual-model}

The fundamental premise of this research is that musical composition is inherently structured and hierarchical.
In traditional music theory, a piece is not merely a sequence of notes, but a complex organization of motifs, phrases,
sections, and movements.
However, many modern digital tools and domain-specific languages treat music as a linear stream of events or a flat
sequence of commands.
This research proposes a shift toward a \textit{structural representation}, where musical ideas are treated as modular,
reusable, and composable units.

\subsection{Linear vs. Structural Representation}\label{detail-subsec:linear-vs.-structural-representation}

In linear representation models, commonly found in live coding environments, the focus is on the \textit{performance}.
Time is often modeled as a continuous thread where commands such as \texttt{play} and \texttt{sleep} are executed
sequentially to generate sound.
While effective for improvisation and real-time interaction, this model often sacrifices the ability to express complex
temporal relationships and hierarchical dependencies.
Modifying a single motif in a linear script often requires broad changes to subsequent timing logic, leading to fragile
and non-modular code.

In contrast, the proposed structural representation models music as a \textit{recursive hierarchy}.
Under this model, a composition is built by nesting smaller musical units within larger ones.
A single note is the base case of this hierarchy.
Notes are grouped into motifs (patterns), motifs are sequenced into phrases, and phrases are assigned to parallel
execution contexts (tracks).
This approach allows for a clean separation of concerns: the musical content (the ``what'') is defined independently of
its temporal placement (the ``when'').

\subsection{Music as a Pure Function}\label{detail-subsec:music-as-a-pure-function}

By treating musical structures as first-class entities, the language allows them to be manipulated using concepts
derived from functional programming.
A musical pattern can be viewed as a \textit{pure function}: it encapsulates a specific musical logic that produces a
deterministic sequence of events.
Just as a function in programming can be called multiple times with different arguments, a musical pattern in
Motivo can be instantiated across different tracks, at different times, or with different parameters (such as
pitch transpositions or rhythmic stretching).

This functional mapping has three consequences for the language design.

It supports modularity because Musical ideas can be defined once and reused throughout a composition.

It supports abstraction because Complex polyphonic structures can be hidden behind simple pattern interfaces.

It supports determinism because Because the language lacks a real-time runtime, every musical relationship is
resolved at compile-time, producing a deterministic representation in the output artifact.

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[
      level/.style={rounded corners=4pt, draw=motivoInk!60, very thick, minimum width=10.5cm, minimum height=0.95cm,
      align=center, font=\sffamily},
      arrow/.style={-Stealth, very thick, draw=motivoInk!70}
    ]
    \node[level, fill=motivoBlue!15] (program) {Motivo program: tempo, time signature, global definitions};
    \node[level, fill=motivoCyan!18, below=0.35cm of program] (track) {Tracks: instrumental identity and MIDI
    channel allocation};
    \node[level, fill=motivoGreen!18, below=0.35cm of track] (voice) {Voices: parallel timeline fragments
    inside a track};
    \node[level, fill=motivoAmber!20, below=0.35cm of voice] (pattern) {Patterns: reusable parameterized
    musical procedures};
    \node[level, fill=motivoRose!14, below=0.35cm of pattern] (values) {Values: notes, rests, chords,
    sequences, drum notes, scalars, booleans};

    \draw[arrow] (program) -- (track);
    \draw[arrow] (track) -- (voice);
    \draw[arrow] (voice) -- (pattern);
    \draw[arrow] (pattern) -- (values);
  \end{tikzpicture}
  \caption{Motivo's source-level hierarchy.
  The compiler validates this hierarchy before lowering it into timestamped note events.}\label{fig:motivo-hierarchy}
\end{figure}
