\section{Imperative Syntax for Algorithmic Variation}\label{sec:syntax-design}

The choice of syntax in a domain-specific language is a critical decision that influences both the learning curve and
the expressiveness of the language.
For Motivo, a C-style imperative syntax was selected for its familiarity and its ability to clearly represent
the deterministic progression of a musical program.

\subsection{Familiarity and Cognitive Load}\label{detail-subsec:familiarity-and-cognitive-load}

By adopting standard programming conventions---such as curly braces for blocks, semicolons for statement
termination, and
parentheses for parameters---Motivo targets a user base that is already familiar with general-purpose languages like C,
C++, or Java.
This choice reduces the cognitive load on the user, allowing them to focus on the musical logic rather than learning a
completely novel syntactical paradigm.
This contrasts with languages like \textit{Tidal}, which rely on deeply nested functional structures that can be opaque
to users without a strong Haskell background.

\subsection{Imperative Flow as a Musical Metaphor}\label{detail-subsec:imperative-flow-as-a-musical-metaphor}

Musical composition is often an imperative process: ``play this, then repeat that, then if a condition is met, change
this''.
Motivo's control flow structures map directly to these compositional needs:

\noindent \textbf{Loops (\texttt{loop}, \texttt{for}):} The \texttt{loop} construct provides a simple mechanism for
repetition (e.g., a constant drum beat).
The C-style \texttt{for} loop enables more complex \textit{algorithmic composition}, where an index variable can
drive musical parameters.
For example, a loop can transpose a motif incrementally by using the iterator value as a pitch offset.

\noindent \textbf{Conditionals (\texttt{if}, \texttt{else}):} These structures allow for \textit{context-aware
composition}.
A phrase might have a different ending depending on whether it is the first or second time it is played, or a track
might include a crash cymbal accent only on the first beat of a section.

\subsection{Algorithmic Variation vs. Static
Definition}\label{detail-subsec:algorithmic-variation-vs.-static-definition}

Traditional MIDI editing is inherently static; a motif is a series of fixed events.
The imperative syntax of Motivo allows motifs to be defined \textit{algorithmically}.
By combining control flow with variables (\texttt{let}), a composer can write code that generates variations of a theme
without manually copying and pasting events.
This methodology reinforces the idea that music is not just a sequence of data points, but a series of logical
instructions that can be computed.
Since the compiler unrolls these instructions at compile-time, the composer enjoys the power of algorithmic generation
without the overhead or latency of a real-time interpreter.

The following compact example shows the intended relationship between imperative notation and musical
structure. The source program contains no MIDI status bytes and no delta-time calculations. It expresses a
bar-level rule, and the compiler expands that rule into concrete events.

\begin{lstlisting}[language=Motivo, caption={A compact Motivo drum pattern.}, label={lst:motivo-groove}]
tempo 128;
signature 4/4;

track percussion using drums {
    pattern bar() {
        for (let i = 0; i < 16; i = i + 1) {
            play hihat from i;
            if (i % 8 == 0) { play kick from i; }
            if (i % 8 == 4) { play snare from i; }
        }
    }

    loop (500) { play bar(); }
}
\end{lstlisting}
