\section{Motivation and Problem}\label{sec:motivation}

In a digital audio workstation, the piano roll is intuitive but brittle: transposing a repeating phrase or
changing its last repetition means editing hundreds of events by hand, and the original algorithmic intent is
lost. Live-coding languages recover structure, but they target low-latency performance and synthesis rather
than a portable offline artifact. General-purpose MIDI scripting offers offline output without musical
abstractions, so composers end up engineering delta-times instead of writing music.

Motivo closes that gap with an offline compiler: the same source always yields the same Format~1 MIDI file, with
no runtime interpreter required at playback. Patterns, tracks, voices, and familiar control flow describe the
piece; the compiler expands them into a timeline of note events.

\subsection{Contrast with Raw MIDI Scripting}\label{detail-subsec:raw-midi-scripting-contrast}

Motivo syntax is defined formally in Chapter~\ref{ch:language-design}; the listing below is intended to be readable
without that context.
For offline composition, many developers use general-purpose programming languages combined with MIDI libraries, such as
Python's \texttt{mido} or C++'s \texttt{RtMidi}.
While these tools offer absolute control over the output, they operate at an exceptionally low level of abstraction.

\topic{Event Placement and Musical Intent.}
When using a library like \texttt{mido}, the composer must manually construct \texttt{Note-On} and \texttt{Note-Off}
messages.
Crucially, because the MIDI file format relies on delta-times, the composer is forced to manually calculate the tick
difference between every sequential event across all tracks.

Motivo abstracts this entirely.
A composer writing \texttt{play (C4, E4, G4):2;} in Motivo communicates their \textit{musical intent} (playing a
C-Major triad for two beats).
Motivo's lowering engine and backend automatically handle the translation into absolute beats, the flattening into
individual notes, the sorting by absolute tick, and the final variable-length delta-time encoding.

\topic{Boilerplate and Expressiveness.}
A simple four-beat loop of a hi-hat playing eighth notes requires a significant amount of boilerplate in
Python: creating
a file, creating a track, appending tempo metadata, and writing a loop that interleaves \texttt{Note-On} and
\texttt{Note-Off} messages with explicit delta times.

Motivo achieves the same result with minimal syntax:
\begin{lstlisting}[language=Motivo, caption={Hi-hat loop in Motivo versus raw MIDI scripting.},
label={lst:midi-scripting-comparison}]
tempo 120;
track using drums {
    loop (8) { play [hihat :0.5]; }
}
\end{lstlisting}

This contrast in verbosity illustrates Motivo's intended role as a domain-specific tool.
By embedding musical semantics (like default durations, rest tracking, and automatic track/channel
assignment) directly into
the language design, Motivo is intended to reduce cognitive friction and let the user operate strictly within the
musical domain.
