\section{Structure of This Thesis}\label{sec:thesis-structure}

The remainder of this thesis follows the path from problem to artifact to evidence.
The present chapter has set out the problem, its motivation and relevance, the representative use cases, and the
contributions that frame the work.

Chapter~\ref{ch:background} establishes the foundations the later chapters build on: it situates the work in its
digital-music context, gives concise primers on music notation and MIDI and on compiler theory, surveys algorithmic
composition and related musical domain-specific languages, and from that survey identifies the research gap that
Motivo addresses.

The two chapters that follow describe the language and the tool that realizes it.
Chapter~\ref{ch:language-design} presents the design of the Motivo language, covering its grammar, type system,
semantics, and temporal model, and tracing the evolution from version~1.0 to version~2.0.
Chapter~\ref{ch:compiler-implementation} then turns to the implementation of \texttt{motivoc}, walking through the
compilation pipeline from the frontend and semantic pass to lowering and the MIDI backend.

Chapter~\ref{ch:studio-engineering} steps outward from the compiler to Motivo Studio and the engineering operations
around it: the web IDE, the server and persistence layer, and the testing, continuous integration, and deployment
practices that maintain it as a product rather than a prototype.

The final chapters return to evidence and reflection.
Chapter~\ref{ch:evaluation} reports the empirical evaluation, measuring compilation time and event expansion on
benchmarks and real compositions and noting the limits of the approach.
Chapter~\ref{ch:conclusion} closes by drawing together the contributions, acknowledging the limitations, outlining
future work, and revisiting the use cases introduced here in light of the evidence gathered along the way.
