\chapter{Introduction}\label{ch:introduction}

Music composition is an inherently structural discipline. Classical sonata forms, jazz chord progressions,
and electronic dance music loops all rely on the hierarchical organization of motifs, phrases, and sections.
Despite this underlying mathematical and logical foundation, modern digital music production relies
predominantly on linear, visual interfaces. While these tools excel at micro-editing, they often lack the
abstractions necessary to manipulate music as structured, algorithmic data.

\section{Motivation}

In a standard Digital Audio Workstation (DAW), the primary method of input is the "Piano Roll"—a grid where
notes are drawn manually across a timeline. This approach is highly intuitive but fundamentally destructive
from a structural perspective. If a composer wishes to transpose a repeating 16-bar melody, change its
rhythm, or conditionally alter its ending on the final repetition, they must manually select and modify
hundreds of individual note events. The logical intent (e.g., "repeat this phrase four times but play the
snare on the off-beat during the last repetition") is lost, replaced by a static collection of data points.

To address this, programmers often turn to Algorithmic Composition. However, existing domain-specific
languages for music (such as Sonic Pi or Tidal) are largely tailored toward \textit{live coding}. They
prioritize real-time performance, utilizing threaded execution models and continuous audio synthesis. While
incredibly powerful for improvisation, these environments introduce complexities related to thread
synchronization, execution latency, and audio engine configuration, which can be overwhelming for composers
whose primary goal is offline, deterministic arrangement.

Furthermore, general-purpose programming languages equipped with MIDI libraries (e.g., Python with
\texttt{mido}) offer offline generation but lack domain-specific abstractions. The composer is burdened with
calculating explicit delta-times and manually constructing low-level binary messages, shifting the focus away
from musical creativity and toward byte-level engineering.

\section{Problem Statement}

There exists a gap in the current ecosystem of music programming tools: a lack of a domain-specific language
optimized for \textit{offline, structural composition} that produces standard, interoperable artifacts.
Specifically, the challenge lies in creating a system that allows composers to define musical ideas as
reusable, parameterized algorithms while guaranteeing that the resulting temporal structure is mathematically
precise, free of real-time execution jitter, and universally readable by standard production software.

\section{Proposed Solution}

This thesis proposes the design and implementation of a novel, compiled Domain-Specific Language (DSL) for
musical composition. The system aims to bridge the gap between algorithmic logic and the rigid binary
specifications of the MIDI protocol.

The proposed solution is built upon the following core principles:
\begin{enumerate}
  \item \textbf{Music as a Recursive Hierarchy:} The language abstracts musical motifs into functions
    (patterns) that can be parameterized, reused, and nested, reflecting the true hierarchical nature of musical form.
  \item \textbf{C-Style Imperative Flow:} By utilizing familiar control flow structures (\texttt{for},
    \texttt{loop}, \texttt{if}), the DSL allows composers to procedurally generate variations and repetitions
    without learning esoteric functional paradigms.
  \item \textbf{Zero-Runtime Compilation:} The compiler operates strictly offline. It performs semantic
    validation and then uses a stateful lowering engine to completely unroll the composition's logic into a
    flat timeline. This guarantees deterministic temporal resolution without the overhead of an interpreter.
  \item \textbf{MIDI as a Semantic Target:} The compiler serializes the final timeline into a Standard MIDI
    File (Format 1). This ensures absolute interoperability, allowing the generated compositions to be
    imported into any DAW for professional orchestration and mixing.
\end{enumerate}

\section{Thesis Structure}

The remainder of this thesis is structured as follows:
\begin{itemize}
  \item \textbf{Chapter 2 (Background):} Provides the foundational knowledge required to understand the
    system, including a primer on music theory, the technical specifications of the MIDI protocol, and the
    basic stages of compiler construction.
  \item \textbf{Chapter 3 (Related Work):} Analyzes existing music programming environments and low-level
    synthesis languages, highlighting the design gaps that motivate this research.
  \item \textbf{Chapter 4 (Methodology):} Details the conceptual framework of the DSL, explaining the mapping
    of musical primitives to programming constructs and the rationale behind the compiler's temporal logic.
  \item \textbf{Chapter 5 (Implementation):} Offers a deep technical dive into the C++23 codebase. It covers
    the Flex/Bison frontend, the recursive semantic analyzer, the stateful lowering engine, and the binary
    MIDI serializer. It also discusses the architecture of the web-based demonstration playground and the
    verification suite.
  \item \textbf{Chapter 6 (Evaluation):} Assesses the performance and utility of the compiler, focusing on
    execution speed, memory footprint, and the expressiveness (compression ratio) of the language compared to
    manual event placement.
  \item \textbf{Chapter 7 (Conclusion):} Summarizes the contributions of the thesis, discusses its current
    limitations, and outlines potential avenues for future research and expansion.
\end{itemize}
