\chapter{Design of the Motivo Language}\label{ch:language-design}

Motivo is built around one design decision: the source program describes musical structure, and the compiler
resolves that structure before playback.
The language therefore separates the composer's symbolic intent from the MIDI event stream that eventually stores the
piece.
This chapter specifies that language: its hierarchical model, formal grammar, type system, temporal semantics, and
compilation strategy at the level of \textit{what} the composer writes and \textit{what} the compiler must guarantee
before code generation begins.

The description reflects \textit{version~2.0}, with explicit static types on variables and pattern parameters;
version~1.0's untyped \texttt{let} bindings and arity-only pattern overloading are summarized only where the evolution
narrative requires contrast.
It also states an important boundary of the current implementation: key-signature declarations are not part of the
language.
Motivo currently accepts tempo and time-signature metadata, while harmonic context remains a future extension.

\input{chapters/02-motivo-language-design/sections/conceptual-model}
\input{chapters/02-motivo-language-design/sections/source-grammar}
\input{chapters/02-motivo-language-design/sections/musical-primitives}
\input{chapters/02-motivo-language-design/sections/syntax-design}
\input{chapters/02-motivo-language-design/sections/semantic-analysis}
\input{chapters/02-motivo-language-design/sections/temporal-model}
\input{chapters/02-motivo-language-design/sections/compilation-strategy}

The language design is intentionally separate from its implementation.
The next chapter describes how \texttt{motivoc} realizes this specification in C++: the compilation pipeline, module
architecture, and the implementation of each compiler stage.
