\chapter{Conclusion}\label{ch:conclusion}

This thesis presented the design, implementation, and evaluation of a novel Domain-Specific Language (DSL)
for musical composition.
By treating musical structures as first-class programming constructs and shifting the complexity of temporal
resolution entirely to a compile-time lowering engine, the project established a new paradigm for offline,
deterministic algorithmic composition.

\section{Summary of Contributions}

The primary contribution of this research is the successful demonstration of a "zero-runtime" architecture
for a musical DSL.
Unlike live-coding environments that rely on threaded execution and real-time interpreters, this compiler
successfully maps high-level musical intent into a static, perfectly quantized Intermediate Representation (IR).

Key achievements include:
\begin{itemize}
  \item \textbf{Structural Abstraction:} The language provides robust constructs (\texttt{pattern},
    \texttt{track}, \texttt{voice}) that allow composers to build complex polyphonic arrangements modularly,
    without manually calculating absolute timestamps or delta-times.
  \item \textbf{Deterministic Lowering:} The implementation of a stateful cursor mutation algorithm
    successfully "unrolls" C-style imperative control flow (loops, conditionals) and recursive pattern calls
    into a linear sequence of MIDI events, while perfectly preserving temporal gaps and rests.
  \item \textbf{Type-Safe Semantics:} The custom semantic analyzer rigorously enforces musical logic,
    ensuring that operators are applied correctly and that patterns are invoked with valid arguments, thereby
    preventing invalid data from reaching the backend.
  \item \textbf{Robust Tooling:} The development of a full-stack, web-based playground with real-time
    diagnostics and Piano Roll visualization validates the compiler's performance and provides an accessible
    environment for users.
\end{itemize}

\section{Limitations}

While the compiler achieves its primary goals, the architectural choices introduce certain limitations:
\begin{itemize}
  \item \textbf{Lack of Real-Time Interaction:} Because the output is fully resolved at compile-time into a
    static file, the language cannot respond to external input (such as live MIDI controllers or network
    signals) during playback.
  \item \textbf{MIDI Expressiveness Constraints:} The current implementation is bound by the constraints of
    the MIDI 1.0 standard. Features like continuous parameter automation (e.g., filter sweeps, pitch bends)
    or microtonal tunings are not currently supported by the AST.
  \item \textbf{Static Pattern Signatures:} Pattern parameters currently only accept concrete evaluated
    types. The language lacks support for passing un-evaluated functions or lambdas, which limits the
    potential for higher-order musical transformations.
\end{itemize}

\section{Future Work}

The foundation laid by this compiler opens several promising avenues for future research and development:

\subsection{Standard Library and Advanced Musical Constructs}
Future iterations could include a standard library of musical utilities. Built-in functions for generating
scales, arpeggios, and chord inversions would greatly enhance the language's expressiveness. Additionally,
extending the type system to understand key signatures and functional harmony (e.g., automatically resolving
Roman numeral chord progressions) would elevate the DSL to a higher level of musical abstraction.

\subsection{Continuous Parameter Automation}
To support more expressive performances, the AST and lowering engine could be extended to support continuous
data. Constructs could be added to define "envelopes" or "curves" over time, which the backend would then
serialize into a dense stream of MIDI Control Change (CC) or Pitch Bend messages.

\subsection{Alternative Backends}
The separation of the IR from the backend makes it feasible to support output formats beyond MIDI. A
compelling future project would be the development of an Open Sound Control (OSC) backend or a direct audio
synthesis backend (e.g., generating Csound or SuperCollider scripts), allowing the language to directly
interface with modern digital signal processing environments.

\vspace{1cm}
In conclusion, this thesis demonstrates that the principles of compiler design and functional abstraction can
be highly effective when applied to musical composition. By bridging the gap between algorithmic logic and
the rigid binary specifications of the MIDI protocol, the resulting DSL empowers composers to explore complex
structural arrangements with precision, confidence, and speed.
