\chapter{Related Work}\label{ch:related-work}

%% #####################################################################################################################

\section{Live Coding and Music Programming Languages}\label{sec:live-coding-and-music-programming-languages}

%% =====================================================================================================================

\subsection{Tidal}\label{subsec:tidal}

\textbf{Tidal} is a domain specific programming language embedded in \textit{Haskell} for encoding musical patterns,
meant to be used during live performances~\cite{McLean_Wiggins_2010}.
It adopts a pattern-based model of music representation, where musical structures are defined as transformations of
time-varying patterns rather than explicit event placement.
It is meant to address the gap in pattern languages, as most languages focused on encoding and decoding musical patterns
at that time were focused more on \textit{musical analysis} rather than \textit{musical
composition}~\cite{McLean_Wiggins_2010}.

%% ---------------------------------------------------------------------------------------------------------------------

\subsubsection{Key Features}

Since it is embedded in the programming language \textit{Haskell}, it inherits most of its syntax and paradigms.
It uses Haskell's strict type system, which in term leads to stricter constraints such as static typing and pure
function usage.
However, it makes use of these constraints to its advantage by leveraging \textit{Haskell's} concepts of
\textit{functors} and \textit{monads}~\cite{McLean_Wiggins_2010}.

\paragraph{Pattern Composition}

Patterns can be composed of sub-patterns; this relation can go on to an arbitrary depth.
This allows for creating complex structures from simple building blocks, creating an organized, \textit{hierarchical}
structure.
Patterns can be concatenated, merged, or even combined by multiplying their numerical patterns, offering great control
and extensibility through pair-wise operations~\cite{McLean_Wiggins_2010}.

\paragraph{Representation and Access}

Some languages represent patterns by using lazily loaded lists, where values are calculated only when they are needed,
which allows for very long or even infinite patterns to be computed and represented.
\textit{Haskell} itself uses lazy lists by default, but \textbf{Tidal} decides to not follow the same idea.

Lazy lists don't work great for live performances since some elements are calculated based on previous items in that
list.
If you modify the definition of a lazy list, in order to continue from a set point, you'd have to recompute
the entire list up to that point, which is computationally expensive, especially in a live coding environment.
This is why \textbf{Tidal} allows \textit{random access} by use of \textit{functions} mapping time values to
events~\cite{McLean_Wiggins_2010}.

\paragraph{Timing}

Rhythmic structures progress linearly, thus it is not that complex to encode these progressions and allow for periodic
or infinite patterns, concepts that \textbf{Tidal} allows.
Timing can be either \textit{discrete} or \textit{continuous}, and musical tradition combines these concepts by often
using discrete notations (like the Western musical staff), paired with playing phrases subtly over continuous time.
\textbf{Tidal} reflects this by defining patterns as \textit{events over discrete time steps}, with the possibility of
defining some floating point onsets as well~\cite{McLean_Wiggins_2010}.

%% ---------------------------------------------------------------------------------------------------------------------

\subsubsection{Limitations}

That said, \textbf{Tidal} also comes with some limitations, such as its reliance on \textit{Haskell's} paradigms and
structure, sacrificing syntax for a higher learning curve, but with the benefit of powerful and efficient pattern
construction.

It also lacks compiling capabilities, as it is meant for outputting events and not actually synthesize sound on its own.
Thus, it relies on \textbf{OSC} (\textit{Open Sound Control}) to stream musical events to an external synthesizer.
This is of course due to its main target being live performances, but this in term makes music composition
for bedroom-producers much harder, as they would have to rely on external hardware~\cite{McLean_Wiggins_2010}.

%% ---------------------------------------------------------------------------------------------------------------------

\subsubsection{Relevance to Our DSL}

\textbf{Tidal} describes how to encode one of the most fundamental constructs in music theory and musical composition.
It shows how we can make pattern declarations more consistent and concise through definitions, rather than hand-placed
notes.

These limitations, however, highlight a different design focus compared to the DSL proposed in this thesis,
which emphasizes deterministic composition and direct compilation to output formats.
