\section{MIDI as a Symbolic Interchange Format}\label{sec:midi-background}

The Musical Instrument Digital Interface (MIDI) is an industry-standard technical protocol developed in 1983
to enable electronic instruments, computers, and related devices to communicate and synchronize with each
other~\cite{Moog_1986}.
Crucially, MIDI does not transmit digital audio (sound waves); instead, it transmits \textit{event messages} that act as
instructions for how music should be produced by a receiving synthesizer.
Digital audio workstations import such messages as editable clips; here we focus on the protocol mechanisms
Motivo's backend must emit.

\subsection{Event-Based Representation}\label{detail-subsec:event-based-representation}

A MIDI stream is composed of sequential commands, whose binary encoding and Standard MIDI File container are defined by
the Standard MIDI Files specification~\cite{MMA_SMF_1996}.
The most fundamental commands are:

\topic{Note-On.} Instructs the synthesizer to begin sounding a pitch.
It carries two data bytes: the \textit{key number} (0--127, where 60 is typically Middle C) and the
\textit{velocity} (0--127, representing how hard the key was struck).

\topic{Note-Off.} Instructs the synthesizer to stop sounding the pitch.

\topic{Program Change.} Changes the active instrument (or ``patch'') on a specific channel (e.g., switching
from a piano to a violin).

Because MIDI transmits events rather than waveforms, a piece of music represented in MIDI is highly editable.
A composer can retroactively change the pitch, duration, or instrument of any note without degrading the audio quality,
making MIDI the \textit{de facto} standard for digital composition~\cite{Moog_1986}.
Moore argues that MIDI conflates musical structure with performance data and lacks native support for higher-level
abstractions such as phrasing and hierarchy~\cite{Moore_1988}.
We therefore treat MIDI as an excellent \textit{backend} target while keeping compositional intent in the source
language, a design choice Motivo adopts deliberately.

\subsection{Channels and General MIDI}\label{detail-subsec:channels-and-general-midi}

To support polyphony across different instruments, MIDI data is transmitted across 16 independent channels.
A synthesizer can be configured to play a bass patch on Channel 1 and a piano patch on Channel 2 simultaneously.
The \textit{General MIDI} (GM) specification standardizes the assignment of instruments to program numbers (e.g.,
Program 1 is Acoustic Grand Piano) and strictly reserves \textbf{Channel 10} for unpitched percussion, where individual
key numbers trigger different drum samples (e.g., Key 36 is the Bass Drum).

\subsection{Delta-Times and Standard MIDI Files (SMF)}\label{detail-subsec:delta-times-and-standard-midi-files-(smf)}

When saving MIDI data to a file (a \texttt{.mid} file), the Standard MIDI File format uses
\textit{delta-times}~\cite{MMA_SMF_1996}.
Instead of recording the absolute time at which an event occurs, a Standard MIDI File records the number of ``ticks'' to
wait after the \textit{previous} event before executing the current one.
Ticks are subdivisions of a quarter note, defined by the file's \textit{Pulses Per Quarter-note} (PPQ) resolution (e.g.,
480 PPQ).
This relative timing system is efficient for serialization but requires careful calculation when translating from
absolute musical beats into a binary file format, a task Motivo's MIDI backend performs during code generation.
