\section{Research Gap and Positioning of Motivo}\label{sec:motivo-position}

The domain-specific languages surveyed above split into two broad families.
Live-coding systems prioritize interactive performance and pattern transformation, streaming events to a synthesis
backend rather than producing a self-contained offline artefact~\cite{Aaron_2016,McLean_Wiggins_2010,Roos_McLean_2023,
Du_Bois_Ribeiro_2019}.
Real-time audio languages offer precise control over synthesis and timing, but at an abstraction level oriented
toward signal graphs and concurrent processes rather than hierarchical song
structure~\cite{Wang_Cook_2003,McCartney_2002}.
None of the surveyed DSLs treats offline, deterministic compilation to a portable symbolic file as its primary design
goal, nor do they combine typed imperative source with a zero-runtime expansion model.

Table~\ref{tab:related-work-comparison} summarizes those languages and the trade-offs that motivate Motivo.

\begin{table}[htbp]
  \centering
  \footnotesize
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{M{2.2cm}M{2.5cm}M{4cm}M{5cm}}
    \toprule
    \textbf{System / Language} & \textbf{Class} & \textbf{Temporal / structural model} & \textbf{Main gap
    relative to this thesis} \\
    \midrule
    Sonic Pi & Live coding & Sequential execution with notes, sleeps, and simple patterns &
    Live synthesis target; limited structural abstraction for offline MIDI compilation \\
    Tidal / Strudel & Live coding & Time-varying patterns with functional composition &
    Event streaming to external backends; no deterministic offline MIDI as primary artifact \\
    HMusic & Live coding & Grid-like patterns and parallel tracks & Compiles to Sonic Pi; remains performance-coupled \\
    ChucK & Real-time synthesis & Explicit time advancement and synchronized concurrent shreds &
    Low-level time management; weak high-level compositional structure \\
    Super\-Collider & Real-time synthesis & UGens, patterns, and streams over OSC &
    Synthesis-oriented; composition and execution not separated \\
    \bottomrule
  \end{tabular}
  \caption{Comparison of the main musical domain-specific languages surveyed in this
  chapter.}\label{tab:related-work-comparison}
\end{table}

From this analysis, three gaps among existing musical DSLs emerge.
First, most target live performance or synthesis, leaving offline deterministic compilation underexplored as a primary
design goal.
Second, few systems bridge low-level audio programming and high-level compositional structure with a zero-runtime
compilation model.
Third, composition and execution are often conflated, so musical intent remains tied to performance-time scheduling.

Motivo addresses these gaps by providing a typed, imperative source language for offline symbolic composition;
a compiler that validates structure and expands it deterministically into a flat event timeline; and a Standard MIDI
File as the portable output artifact.
Patterns and loops compress repetitive structure in the source program while the lowering stage materializes the full
event stream---the compression argument introduced above.
