\section{Compiler and Domain-Specific Language Foundations}\label{sec:compiler-dsl-foundations}

A compiler is a specialized software system that translates source code written in a high-level programming language
into a lower-level representation, such as machine code, bytecode, or---in the context of this thesis---a binary MIDI
file.
The compilation process is traditionally divided into several distinct phases to manage complexity and support
systematic validation~\cite{Aho_Lam_Sethi_Ullman_2006,Cooper_Torczon_2011}.

\subsection{Lexical and Syntactic Analysis}\label{detail-subsec:lexical-and-syntactic-analysis}

The first phase, \textit{lexical analysis} (scanning), reads the raw character stream of the source file and groups
characters into meaningful sequences called \textit{tokens}.
For example, the string \texttt{tempo} might be converted into a keyword token, while \texttt{120} becomes an integer
literal token.
Regular expressions and finite automata provide the theoretical basis for recognizing token classes; in practice,
scanner generators automate this phase from a token specification~\cite{Aho_Lam_Sethi_Ullman_2006}.

The second phase, \textit{syntactic analysis} (parsing), takes the stream of tokens and verifies that they form a
valid sequence according to the formal grammar of the language.
Context-free grammars and pushdown automata model this step~\cite{Aho_Lam_Sethi_Ullman_2006}.
The parser organizes the flat token stream into a hierarchical tree structure known as an \textit{Abstract Syntax Tree}
(AST).
The AST captures the logical operations of the code (e.g., nesting a \texttt{play} statement inside a \texttt{loop}
block) while discarding irrelevant textual details like whitespace and semicolons.

\subsection{Semantic Analysis}\label{detail-subsec:semantic-analysis}

While the parser ensures that a sentence is grammatically correct, the \textit{semantic analyzer} ensures that it makes
logical sense.
This phase traverses the AST to perform checks that cannot be captured by a context-free
grammar~\cite{Cooper_Torczon_2011}.
The semantic phase therefore centers on two responsibilities.

\noindent \textbf{Type checking.} Ensuring that operations are applied to compatible data types (e.g., preventing the
addition of a boolean to a note).

\noindent \textbf{Symbol resolution.} Maintaining a \textit{symbol table} to ensure that variables and functions are
declared before they are used, and that they are accessed within their valid scope.

\subsection{Intermediate Representation and Code
Generation}\label{detail-subsec:intermediate-representation-and-code-generation}

After validation, the compiler often translates the AST into an \textit{Intermediate Representation} (IR).
An IR is a lower-level, often flattened representation of the program that is easier to optimize and translate than the
hierarchical AST~\cite{Cooper_Torczon_2011}.
In standard compilers this might involve generating three-address code.
In the context of music generation, lowering to an IR involves unrolling loops and resolving relative timings into
absolute timestamps.

Finally, the \textit{code generation} phase translates the IR into the target output format.
This phase handles the intricate binary serialization details, such as endianness encoding and protocol-specific
constraints, yielding the final artifact.

\subsection{The Motivo Compilation Pipeline}\label{subsec:motivo-pipeline}

Motivo follows the phased structure described above.
The end-to-end pipeline implemented in this thesis is
\texttt{parse} $\rightarrow$ \texttt{semantic analysis} $\rightarrow$ \texttt{lowering} $\rightarrow$
\texttt{MIDI code generation}.

Lexical and syntactic analysis produce an AST from Motivo source.
Semantic analysis performs a single traversal that validates types, resolves pattern overloads by parameter signature,
and attaches type annotations to expressions.
Lowering expands control flow, pattern calls, and track structure into a flat intermediate program of
timestamped musical
events.
The MIDI backend serializes that program into a Standard MIDI File.
There is no separate optimization pass: the compiler's role is validation and deterministic expansion rather than
runtime scheduling.

Domain-specific languages are useful when a problem has stable concepts that are awkward to express directly
in a general-purpose language. Van Deursen, Klint, and Visser describe DSLs as languages that trade
generality for concise notations and domain-level abstractions~\cite{van_Deursen_Klint_Visser_2000}. Fowler
makes a similar software-engineering argument: a DSL can be worthwhile when its notation lets domain ideas be
expressed more directly than the host implementation language~\cite{Fowler_2010}. Motivo applies that
argument to symbolic composition: the implementation is written in C++23, but the user writes musical
structures rather than C++ vectors of binary MIDI messages.

\subsection{Structure of This Thesis}\label{subsec:thesis-roadmap}

The remainder of this document follows the Motivo system from motivation through evaluation.
Table~\ref{tab:thesis-roadmap} summarizes what each chapter covers.

\begin{table}[htbp]
  \centering
  \footnotesize
  \renewcommand{\arraystretch}{1.4}
  \begin{tabular}{p{1.2cm}p{4.5cm}p{7.5cm}}
    \toprule
    \textbf{Ch.} & \textbf{Title} & \textbf{What the reader learns} \\
    \midrule
    1 & Research context and theoretical foundations &
    Digital music workflows, MIDI background, compiler theory, related musical languages, and the research gap Motivo
    addresses \\
    2 & Design of the Motivo language and compiler &
    Language grammar, type system, semantic rules, compilation pipeline, and implementation of each compiler stage \\
    3 & System architecture and software engineering &
    Motivo Studio web application: editor, compile API, visualization, playback, deployment, testing, and CI/CD \\
    4 & Empirical evaluation &
    Benchmark methodology, compilation-time measurements, event-expansion behavior, test evidence, and current limits \\
    5 & Conclusions and future directions &
    Contributions, limitations, and directions for extending the language and toolchain \\
    \bottomrule
  \end{tabular}
  \caption{Roadmap of the thesis chapters.}\label{tab:thesis-roadmap}
\end{table}

Chapter~\ref{ch:language-compiler} is the technical core: it explains both the language design and the compiler
implementation in detail.
Chapter~\ref{ch:studio-engineering} treats the web studio as an integral engineering artifact rather than a demo
wrapper.
Chapter~\ref{ch:evaluation} evaluates properties that can be measured from the implementation---compilation speed,
event expansion, and automated test coverage---without claiming to assess musical creativity.
Chapter~\ref{ch:conclusion} summarizes contributions honestly, including known limitations of the current language and
toolchain.
