\section{Compilation Strategy}\label{sec:compilation-strategy}

Motivo adopts an explicitly offline, \textit{zero-runtime} strategy.
Source programs are compositional specifications, not executable performance scripts:
loops, conditionals, and pattern calls are resolved during compilation so the output artifact contains only concrete
timestamped events.

This decision addresses the tension in Section~\ref{sec:motivation} between live-coding immediacy and
DAW-style repeatability.
Live systems privilege on-the-fly interaction but can defer correctness and timing behavior to runtime state.
DAW workflows privilege deterministic playback and editable structure but often lack compact algorithmic abstractions.
Motivo commits to the deterministic side of that spectrum while retaining algorithmic source constructs through
compile-time expansion.

Semantics therefore precede lowering as a hard gate.
Type mismatches, illegal writes across scope boundaries, and invalid play-target combinations are rejected before event
generation begins.
Once validated, lowering reconstructs a flat event timeline that is serialized by the MIDI backend.
Implementation details of this pipeline and its data artifacts are discussed in
Chapter~\ref{ch:compiler-implementation}; the present section focuses only on the language-level strategy.

The zero-runtime commitment is what lets algorithmic intent survive lowering to a flat timeline.
For a fixed source file and compiler version, the generated event schedule is reproducible and independent of
host timing,
thread scheduling, or interactive interpreter latency; the rules remain readable in source even after expansion.

This chapter has specified \emph{what} Motivo is at the language level: its conceptual model, grammar, type system,
temporal semantics, and offline compilation strategy.
The next chapter turns to \emph{how} that specification is realized, walking through \texttt{motivoc} stage by stage,
from source text to Standard MIDI File bytes.
