\section{Types and Compositional Constructs}\label{sec:musical-primitives}

Motivo treats musical material as typed source objects rather than ad hoc event emissions.
This design supports static diagnostics while preserving a compositional vocabulary familiar to musicians:
motifs can be named, reused, and transformed without manually recalculating absolute timestamps.

\subsection{Musical entities as typed values}

The language separates declarable value types from expression-only musical constants.

\topic{Notes.}
Literals such as \texttt{C4} or \texttt{F\#2} denote pitch class plus octave and can be stored in \texttt{note}
variables.

\topic{Chords.}
Parenthesized collections such as \texttt{(C4, E4, G4)} represent simultaneous events.
Optional per-member durations affect local span, and chord duration is interpreted as the maximum member duration.

\topic{Sequences.}
Bracketed lists such as \texttt{[C4, D4:2, rest]} encode ordered cells that may contain notes, chords, and rests.
The sequence type (\texttt{seq}) therefore captures phrase-level structure directly in source code.

\topic{Rests and drum hits.}
\texttt{rest} and drum literals such as \texttt{kick}, \texttt{snare}, and \texttt{hihat} are expression-only symbolic
constants.
They can be scheduled by \texttt{play} and embedded in larger expressions (for example in sequences), but they are not
declaration types.
This preserves a clear distinction between stored musical variables (\texttt{note}, \texttt{chord}, \texttt{seq}) and
domain tokens that represent silence or percussion events.

\topic{Scalars.}
\texttt{int}, \texttt{double}, and \texttt{bool} support algorithmic composition through loop bounds, guards, and
numeric transformations applied to musical expressions.

\subsection{Compositional abstractions}

Three structural constructs organize the hierarchy shown in Figure~\ref{fig:motivo-hierarchy}: patterns for reuse,
tracks for instrumental identity, and voices for controlled parallelism.

\topic{Patterns as reusable units.}
A \texttt{pattern} is a named, typed compositional unit invoked through \texttt{play}.
Patterns can appear globally or inside enclosing track/voice scopes, and lexical shadowing applies in the usual
inside-out manner.
Version~2 overload resolution is signature-based: patterns with the same name may coexist when their parameter type
lists differ (for example, \texttt{fill(bool)} and \texttt{fill(seq)}).
Arguments are evaluated in caller scope before parameter binding, after which the pattern executes in its own local
scope.

\topic{Tracks as instrument and channel contexts.}
Each \texttt{track} defines a primary sequential timeline that lowers to one MIDI channel in the output artifact.
The optional \texttt{using} clause selects instrument identity from a closed keyword set:
\texttt{piano}, \texttt{guitar}, \texttt{bass}, \texttt{violin}, and \texttt{drums}.
This closed vocabulary is a deliberate General MIDI tradeoff: it limits source-level naming freedom but ensures that
instrument semantics are stable and immediately verifiable at compile time.

\topic{Voices as parallel lines inside a track.}
A \texttt{voice} creates a parallel line that inherits the enclosing track cursor (or an explicit \texttt{from}
offset), then advances independently.
This allows contrapuntal layering within one instrumental context without forcing the outer track timeline to wait.
Voice-local declarations remain encapsulated, preserving modularity and avoiding accidental state coupling across
parallel lines.
