\section{Semantic Analysis}\label{sec:semantic-rules}

Semantic analysis establishes that a parsed Motivo program is not only syntactically valid but compositionally
well-formed.
The pass verifies scope visibility, assignment legality, type safety, and instrument-specific musical constraints, then
stores typed bindings for later lowering.

\subsection{Single-pass analysis and design rationale}

Version~2 semantic checking is intentionally organized as one coherent traversal.
When a scope opens, visible pattern signatures are registered; statements and expressions are then validated in source
order within that same walk.
This differs from version~1, where arity-based pattern calls required an additional pass to simulate pattern bodies
under concrete arguments.
Explicit source types in version~2 make contracts visible at declaration sites, so call checking can be performed at the
call itself instead of being deferred.

The practical consequence is earlier feedback:
type and binding errors are reported where the mismatch is introduced, and lowering can consume analysis
results directly
without re-deriving symbol or expression metadata.

\subsection{Scoping model and writable boundaries}

Motivo uses lexical scope stacking (global, track, voice, pattern, and control-flow blocks), but write permissions are
governed by an additional \emph{writable boundary} relation that reflects temporal execution context.

Let $\Gamma$ be the visible symbol environment and $W$ the innermost scope that remains writable from the current
position.
Identifier lookup follows standard lexical rules, but assignment must satisfy both visibility and boundary permissions:
\[
  \Gamma \vdash x : \tau,\quad
  \Gamma \vdash e : \tau,\quad
  \mathrm{scope}(x) \preceq W
  \;\Rightarrow\;
  \Gamma, W \vdash (x \leftarrow e)\ \text{is valid}
\]
If the declaration scope of $x$ lies outside the writable boundary $W$, assignment is rejected even when the type
matches.

\topic{Sequential contexts.}
Track bodies, nested patterns in the same sequential context, and control-flow blocks (\texttt{for}, \texttt{loop},
\texttt{if}) inherit the parent writable boundary.
Because these constructs execute on one timeline, updating shared counters or phrase parameters is deterministic.

\topic{Parallel contexts.}
A \texttt{voice} introduces a new writable boundary at voice scope.
Inside that voice, outer bindings remain readable but are not writable.
This prevents two parallel lines from racing on the same mutable track-level state while still allowing local voice
variables to evolve.

\subsection{Type system rules}

Declarable source types are \texttt{int}, \texttt{double}, \texttt{bool}, \texttt{note}, \texttt{seq}, and
\texttt{chord}.
Declarations and assignments require exact type equality:
\[
  \Gamma \vdash e : \tau
  \;\Rightarrow\;
  \Gamma \vdash \text{declare}(x:\tau, e)\ \text{is valid}
\]
and any $\tau' \neq \tau$ is a type error.
Thus \texttt{int x = 1.0;} is invalid even though numeric operators may produce promoted intermediate results.

Operator families are validated by operand class:
arithmetic operators require numeric operands, logical operators require booleans, relational operators require
compatible comparands, and sequence concatenation is not defined through \texttt{+}.

\subsection{Pattern signatures, overloads, and \texttt{play}}

Pattern resolution is signature-based.
For call-site argument types $\langle \tau_1,\dots,\tau_n \rangle$, the analyzer selects the best-scoring visible
pattern whose parameters are each assignable from the corresponding argument type, using exact type matches as the
tiebreak; since assignability is currently strict type equality, this reduces to an exact signature match in practice.
Name equality and arity alone are insufficient; this enables semantically distinct overloads such as rhythmic and
melodic variants under one pattern name.

\texttt{play} then applies domain constraints:
melodic tracks accept note/chord/sequence-style material, drum tracks accept drum-note material, and optional
\texttt{:duration} and \texttt{from} expressions must type-check as numeric.
Together, these rules replace version~1's arity-only dispatch and second-pass instantiation checks while keeping the
analysis model fully static.
