\section{Source Syntax and Grammar}\label{sec:source-grammar}

Motivo is defined by a lexical scanner and a context-free parser; the authoritative specification is a documented EBNF
grammar.
The scanner recognizes note literals (\verb|[A-G][#b]?[0-8]|), drum names, keywords for structure
(\texttt{track}, \texttt{pattern}, \texttt{voice}), control flow, and declarable types (\texttt{int},
\texttt{double}, \texttt{bool}, \texttt{note}, \texttt{seq}, \texttt{chord}).
\texttt{rest} and drum literals appear in expressions but are not declaration types.

\subsection{Lexical and syntactic strata}

The lexical layer reserves two relevant closed vocabularies.
First, instrument keywords are fixed to \texttt{piano}, \texttt{guitar}, \texttt{bass}, \texttt{violin}, and
\texttt{drums}.
Second, drum-note tokens are fixed to \texttt{kick}, \texttt{snare}, \texttt{hihat}, \texttt{crash}, and
\texttt{ride}.
This bounded token set intentionally privileges deterministic static validation over open-ended runtime patch naming.

At parse level, a program starts with optional header declarations (\texttt{tempo}, \texttt{signature}), followed by
top-level variable declarations, pattern definitions, and track definitions.
Tracks may contain statements, local patterns, and \texttt{voice} blocks; a voice can contain statements and local
patterns but cannot introduce nested tracks, preserving one instrument context per track.

Declarations follow \texttt{<type> IDENT = <expr>;} across global and local scopes.
Pattern parameters use the same type keywords, ensuring one type vocabulary across binding and invocation sites.
\texttt{play} targets notes, rests, chords, sequences, drum literals, identifiers, and pattern calls, with optional
\texttt{:duration} and \texttt{from} modifiers constrained by grammar category.

\subsection{Version 1 to version 2 grammar evolution}
\label{detail-subsec:v1-v2-grammar-evolution}

The transition from version~1 to version~2 is a language-design revision, not only a parser refactor.
Version~1 favored a scripting style with untyped \texttt{let} declarations and arity-only pattern dispatch.
That surface brevity shifted semantic burden to later analysis stages, including a follow-up pass to detect call-shape
inconsistencies that were not explicit in source.

Version~2 replaces \texttt{let} with explicit type-prefixed declarations and typed pattern parameters.
As a result, overload resolution is defined on full parameter type signatures rather than argument count alone.
Because contracts are now visible in grammar-level forms, semantic checking can validate calls in one coherent pass.
The evolution also standardizes program headers: optional \texttt{tempo} and \texttt{signature} declarations precede
top-level definitions, and all compositional structure lives in typed tracks, patterns, and statements below.
