\section{Timeline Semantics}\label{sec:temporal-model}

Motivo schedules all musical events through a cursor model measured in beats.
Rather than relying on runtime clocks, each compositional context carries an abstract playhead whose movement is defined
purely by source semantics.

\subsection{Cursor movement in sequential contexts}

\topic{Relative placement.}
A \texttt{play} statement without \texttt{from} emits material at the current cursor, then advances it by the emitted
duration.
For primitive values, this duration follows the value kind:
notes and rests use their explicit or default duration, chords advance by the longest member span, and sequences advance
by the sum of their flattened members.
The colon modifier sets an explicit duration expression (for example \texttt{play C4:0.5}).

\topic{Absolute placement.}
\texttt{from} schedules material at an absolute beat without moving the local cursor (\texttt{play A4 from 16;}).
This separation allows anchoring events to fixed timeline points while preserving sequential authoring flow around them.

\subsection{Pattern footprints and caller advancement}

Pattern execution uses an internal cursor that starts at zero for each call.
When the call returns, the caller cursor advances by the pattern's \emph{footprint}: the total span reached by that
internal cursor, including embedded rests and control-flow expansion.
This makes pattern calls compositional in time: callers can treat a pattern as one durational unit without inspecting its
internal statement graph.

\subsection{Voice isolation and parallel timelines}

A \texttt{voice} inherits the enclosing track cursor, or a beat specified by \texttt{from}, and then advances on its own
independent cursor.
Advancement inside that voice does not mutate the track cursor and does not mutate sibling voice cursors.
This isolation is essential for deterministic parallelism:
multiple lines can begin from related reference beats yet diverge rhythmically without hidden interference.

After lowering, events from track and voice contexts are merged by absolute beat into the same track output stream.
Consequently, statements that follow a voice declaration in source continue from the track cursor rather than waiting for
the longest voice branch to terminate.

\begin{figure}[htbp]
  \centering
  \motivoscalefig{%
    \begin{tikzpicture}[
        beat/.style={motivo/beat},
        note/.style={rounded corners=2pt, minimum height=0.4cm, draw=white, line width=0.8pt},
        arrow/.style={-Stealth, thick, draw=motivoInk!70}
      ]
      \draw[thick, draw=motivoInk!70] (0,0) -- (11.5,0);
      \foreach \x/\b in {0/0,2/1,4/2,6/3,8/4,10/5,11.5/6} {
        \draw[draw=motivoInk!45] (\x,0.12) -- (\x,-0.12);
        \node[beat, below=0.16cm] at (\x,0) {\b};
      }

      \node[note, fill=motivoBlue!70, minimum width=1.7cm] at (1,0.6) {\textcolor{white}{\scriptsize play C4}};
      \node[note, fill=motivoGreen!70, minimum width=1.7cm] at (3,0.6) {\textcolor{white}{\scriptsize play D4}};
      \node[note, fill=motivoAmber!85, minimum width=1.7cm] at (5,0.6) {\textcolor{white}{\scriptsize play E4}};
      \node[note, fill=motivoRose!80, minimum width=2.0cm] at (8,1.35) {\textcolor{white}{\scriptsize play G4 from 4}};

      \draw[arrow] (0,1.05) -- node[beat, above, align=center] {sequential \texttt{play}\\advances cursor} (6.2,1.05);
      \draw[dashed, thick, draw=motivoInk!55] (8,0.9) -- (8,0.5);
      \node[beat, align=center] at (11.2,1.55) {\texttt{from 4}: beat~4 placement,\\cursor unchanged};
    \end{tikzpicture}%
  }
  \caption{Cursor-based temporal placement.
    Sequential \texttt{play} statements advance the cursor; \texttt{from} places an event at an explicit beat without
  moving the cursor.}\label{fig:cursor-model}
\end{figure}
