\chapter{Related Work}\label{ch:related-work}

%% ---------------------------------------------------------------------------------------------------------------------

\section{Introduction}\label{sec:introduction}

This chapter surveys existing approaches to music representation and programming, focusing on systems that enable the
creation, manipulation, and execution of musical structures through textual or programmatic means.

The reviewed works span multiple paradigms, including live coding environments, real-time audio programming
languages, and textual music representation systems.
These systems differ significantly in their design goals, levels of abstraction, and approaches to modeling musical
time and structure.

Live coding and pattern-based systems, such as \textbf{Sonic Pi}, \textbf{Tidal}, and \textbf{HMusic}, emphasize
expressiveness and real-time interaction, often prioritizing performance and accessibility.

In contrast, real-time audio programming languages like \textbf{ChucK} and \textbf{SuperCollider} provide
fine-grained control over sound synthesis and timing, but operate at a lower level of abstraction.

Additionally, representation-focused systems such as \textbf{MIDI} and \textbf{ABC notation} define how musical
data can be encoded, stored, and transmitted, often prioritizing interoperability and readability over
compositional abstraction.

By analyzing these systems, this chapter identifies key design trade-offs and limitations in existing
approaches, particularly with respect to temporal modeling, abstraction level, and compositional structure.
These observations serve as the foundation for motivating the design of the domain-specific language proposed in
this thesis.

%% ---------------------------------------------------------------------------------------------------------------------

\section{Live Coding and Pattern-Based Systems}\label{sec:live-coding-and-pattern-based-systems}

\input{chapters/03-related-work/sonic_pi}
\input{chapters/03-related-work/tidal}
\input{chapters/03-related-work/hmusic}

%% ---------------------------------------------------------------------------------------------------------------------

\section{Real-Time Audio Programming Languages}\label{sec:real-time-audio-programming-languages}

\input{chapters/03-related-work/chuck}
\input{chapters/03-related-work/super_collider}

%% ---------------------------------------------------------------------------------------------------------------------

\section{Representation}\label{sec:representation}

\input{chapters/03-related-work/midi}
\input{chapters/03-related-work/abc_notation}

%% ---------------------------------------------------------------------------------------------------------------------

\section{Summary and Research Gaps}\label{sec:summary-and-research-gaps}

The systems discussed in this chapter illustrate a wide spectrum of approaches to music programming and
representation, each addressing different aspects of musical creation.
Live coding systems such as \textbf{Sonic Pi} and \textbf{Tidal} prioritize expressiveness and real-time
interaction, enabling rapid experimentation and performance-oriented workflows.
However, they often rely on sequential or pattern-based models that can limit explicit control over complex
temporal relationships.

Real-time audio programming languages, including \textbf{ChucK} and \textbf{SuperCollider}, provide precise
control over timing and sound synthesis.
While these systems offer high flexibility, they operate at a relatively low level of abstraction, requiring
users to manage details related to signal processing, concurrency, and timing explicitly.

Higher-level domain-specific approaches such as \textbf{HMusic} demonstrate how musical structures can be
represented through patterns and tracks, enabling more structured composition.
However, these systems remain primarily pattern-centric and often depend on external backends for execution.

Representation systems such as \textbf{MIDI} and \textbf{ABC notation} focus on encoding musical information for
storage, transmission, and playback.
While widely adopted, these formats lack support for higher-level compositional abstractions and do not provide
mechanisms for structuring music beyond sequences of events or symbolic notation.

From this analysis, several key limitations emerge.
First, many systems lack an explicit and flexible model for representing time, often relying on implicit
sequencing or pattern repetition.
Second, there is a gap between low-level audio programming languages and high-level compositional tools, with
few systems providing both expressive power and structured abstraction.
Third, existing approaches often conflate composition with execution, making it difficult to separate musical
intent from performance details.

These limitations motivate the development of the domain-specific language proposed in this thesis, which aims
to provide a timeline-based model for music composition, support deterministic and structured representations of
musical ideas, and enable modular and extensible compilation to multiple output formats.
