\section{Future Work}\label{sec:future-work}

The limitations in Section~\ref{sec:limitations} suggest a natural roadmap.
Each item below extends Motivo without contradicting the offline, MIDI-first design, but none is implemented in the
present thesis.

\subsection{Richer MIDI expression}\label{subsec:fw-midi-expression}

The most direct gap relative to modern production workflows is expressive control at the note and track level.
Per-note velocity syntax, velocity curves driven by loops or pattern parameters, and MIDI continuous controllers
(including volume and filter-related CC data) would require AST and type-system extensions, IR fields beyond the
current four-tuple model, backend serialization changes, and corresponding updates to the piano-roll visualizer and
Tone.js playback scheduler.
Pitch bend and automation-style envelopes would follow the same pipeline.
The existing separation between IR and backend makes such extensions architecturally plausible, but they remain
substantial cross-cutting work.

\subsection{Modularity, instruments, and libraries}\label{subsec:fw-modularity}

Cross-file \texttt{import} or module boundaries would let composers share pattern libraries across projects without
copying source text.
That feature touches parsing (multi-unit compilation), semantic visibility rules, Studio file relationships, and
possibly a package or namespace convention.
In parallel, a less rigid instrument model---custom GM program numbers, named sound presets, or project-level
instrument tables---would reduce dependence on the current keyword enum.
A standard library of reusable scales, chord voicings, arpeggiators, Euclidean rhythms, and drum templates could sit
atop either extension once harmonic context and module boundaries are defined.

\subsection{Harmonic context}\label{subsec:fw-harmonic-context}

Key-signature declarations would let Motivo reason about scale degrees, diatonic transposition, and chord functions
instead of treating every pitch as an independent literal.
Introducing them carefully requires syntax, semantic interactions with accidentals, and explicit lowering rules so
that deterministic MIDI output remains predictable.
Because key signatures are not part of the current implementation, they belong here rather than in the language
description chapters.

\subsection{Studio product and evaluation depth}\label{subsec:fw-studio-eval}

On the application side, deeper project workflows---multi-file workspaces, version comparison, richer export metadata,
and compile feedback surfaced in the UI (event counts, compile time, density warnings)---would address friction noted
in Section~\ref{subsec:lim-daw-workflow} without changing the compiler's core model.
Alternative backends (JSON event streams, MusicXML-oriented export, or direct synthesis targets) could reuse
the same IR.

Evaluation could grow beyond synthetic throughput benchmarks.
Configurable resource limits replacing the fixed 20,000-event cap, multi-machine performance studies, and user-centered
experiments comparing Motivo to DAW note entry for specific task types would strengthen claims about workflow fit.
A third compiler generation benchmark deferred by the author until a new language feature lands would complement the
existing version~1.0 versus version~2.0 comparison.
