ABSTRACT

\vspace{0.5cm}
\hrule
\vspace{0.5cm}

Digital music production combines creative work with increasingly technical tooling.
Digital audio workstations make detailed editing practical, while textual and live-coding systems show that musical
material can also be described as executable structure.
However, many production workflows still require a composer to manage repeated note events manually, and many music
programming environments focus on real-time performance rather than offline generation of an interoperable score.

This thesis presents Motivo, a domain-specific language and compiler for structured, algorithmic music composition.
Motivo represents musical ideas through notes, rests, chords, sequences, patterns, tracks, and voices.
The language uses familiar imperative control flow, but the compiler resolves that control flow before playback by
lowering the program into an intermediate representation of timestamped musical events.
The current implementation targets Standard MIDI Files, so the result can be inspected, played back, or imported into
other music tools that understand MIDI\@.

The work has two complementary parts.
The first part is the Motivo compiler, implemented in C++23 with a Flex/Bison frontend, semantic analysis, a lowering
stage based on a cursor model, and a MIDI writer.
The second part is Motivo Studio, a full-stack web application that provides a Monaco-based editor, compile diagnostics,
MIDI parsing, piano-roll visualization, browser playback, Docker-based deployment, and continuous integration workflows.

The evaluation combines implementation inspection with local empirical measurements.
On an Apple M3 Pro system, a Release build of \texttt{motivoc} compiled synthetic programs up to the current
20,000-note-event safety limit in approximately 6 ms, including MIDI file writing.
The same benchmark corpus shows the intended source-to-event expansion: a 7-line program can generate 20,000 MIDI note
events, while an 18-line drum groove can generate 10,000 note events.
These results do not claim artistic quality or general superiority over existing tools; they support the narrower claim
that Motivo can express repeated symbolic musical structures compactly and compile them quickly in the tested setting.
