\section{Testing and Continuous Integration}\label{sec:testing-ci}

Quality for Motivo Studio is enforced at four boundaries: the C++ compiler (CTest/GoogleTest), the Express API
(Vitest with optional Postgres integration), the Next.js client (Vitest and Testing Library), and this thesis PDF
(GitHub Actions on \texttt{paper/}).
Counts below were collected on 2026-06-01 at monorepo commit \texttt{7e8a55fb} unless noted.

\subsection{Verified test inventory}\label{subsec:test-inventory}

\begin{table}[htbp]
  \centering
  \renewcommand{\arraystretch}{1.3}
  \begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.8cm}>{\raggedright\arraybackslash}p{2.4cm}X}
    \toprule
    \textbf{Subsystem} & \textbf{Tooling} & \textbf{Counts} & \textbf{Coverage / notes} \\
    \midrule
    Compiler & CTest, GoogleTest & 53 test translation units; \textbf{635} registered CTest cases; 4 golden
    \texttt{.motivo} fixtures in \texttt{tests/golden/fixtures/} & Local Debug run: \textbf{635/635 passed} after
    correcting \texttt{lowering\_errors.motivo} (division semantics no longer triggers lowering-stage errors).
    CI runs \texttt{make ci-test} with gcovr coverage enabled. \\
    Server & Vitest, Supertest & 7 test files; \textbf{31} test cases (\textbf{20} unit without DB; \textbf{11}
    integration skipped locally without Postgres) & Local unit coverage: \textbf{34.6\%} statements,
    \textbf{34.4\%} lines. CI uses Postgres~16 service container and \texttt{npm run test:ci}. \\
    Client & Vitest, Testing Library & 26 test files; \textbf{79} tests (all passed locally) & Local coverage:
    \textbf{49.7\%} statements, \textbf{51.4\%} lines. CI runs \texttt{npm run test:ci}. \\
    Paper & latexmk, tex-fmt, chktex, codespell & 4 jobs on \texttt{paper/**} pushes & Reproducible PDF artifact; no
    coverage metric. \\
    \bottomrule
  \end{tabularx}
  \caption{Test and coverage baseline for thesis claims (2026-06-01).}\label{tab:test-counts}
\end{table}

Compiler unit suites exercise parsing, explicit type rules, lowering, MIDI writing, and integration through
\texttt{motivo::compile()}.
Golden tests compile fixture sources (\texttt{happy\_path.motivo}, \texttt{parse\_errors.motivo},
\texttt{semantic\_errors.motivo}, \texttt{lowering\_errors.motivo}) and assert MIDI shape or collected diagnostics per
stage---not byte-identical comparison to a checked-in \texttt{fur\_elise} binary (that composition ships as a studio
example only).

Server tests cover Zod validation, compile controller success and failure paths with fake \texttt{motivoc}
executables, stderr diagnostic parsing, auth session cookies, and file CRUD against Postgres when available.
Client tests cover compile hooks, marker mapping, MIDI parsing, piano-roll layout hooks, playback utilities, auth
context, file workspace/autosave scheduling, and IDE shortcuts.

\subsection{Monorepo CI workflows}\label{subsec:ci-workflows}

GitHub Actions workflows live at the monorepo root (paths under \texttt{code/} and \texttt{paper/}).

\begin{table}[htbp]
  \centering
  \renewcommand{\arraystretch}{1.25}
  \begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{2.6cm}>{\raggedright\arraybackslash}p{3.2cm}X}
    \toprule
    \textbf{Workflow} & \textbf{Trigger paths} & \textbf{Job progression} \\
    \midrule
    \texttt{compiler.yml} & \texttt{code/compiler/**} & static (clang-format) $\rightarrow$ build (CMake/Ninja)
    $\rightarrow$ test (CTest + gcovr report) $\rightarrow$ package (Docker/build artifacts). \\
    \texttt{server.yml} & \texttt{code/server/**} & lint (ESLint) $\rightarrow$ typecheck $\rightarrow$ format-check
    $\rightarrow$ test (Vitest + Postgres) $\rightarrow$ build $\rightarrow$ package. \\
    \texttt{client.yml} & \texttt{code/client/**} & lint $\rightarrow$ typecheck $\rightarrow$ format-check
    $\rightarrow$
    test $\rightarrow$ build $\rightarrow$ package. \\
    \texttt{paper.yml} & \texttt{paper/**} (excludes \texttt{notes/}, \texttt{scripts/}) & format-check (tex-fmt)
    $\rightarrow$ lint-check (chktex) $\rightarrow$ spell-check (codespell) $\rightarrow$ build-paper (TeX Live PDF). \\
    \bottomrule
  \end{tabularx}
  \caption{CI workflows and their primary stages.}\label{tab:ci-workflows}
\end{table}

\begin{figure}[htbp]
  \centering
  \motivoscalefig{%
    \begin{tikzpicture}[
        job/.style={motivo/mod, minimum width=1.7cm, minimum height=0.72cm},
        arrow/.style={-Stealth, thick, draw=motivoInk!70}
      ]
      \node[job, fill=motivoBlue!14] (format) {static / lint};
      \node[job, fill=motivoCyan!14, right=0.18cm of format] (lint) {format};
      \node[job, fill=motivoGreen!14, right=0.18cm of lint] (types) {build};
      \node[job, fill=motivoAmber!20, right=0.18cm of types] (test) {test};
      \node[job, fill=motivoRose!12, right=0.18cm of test] (package) {package};

      \draw[arrow] (format) -- (lint);
      \draw[arrow] (lint) -- (types);
      \draw[arrow] (types) -- (test);
      \draw[arrow] (test) -- (package);
    \end{tikzpicture}%
  }
  \caption{Typical CI progression per module (exact job names vary slightly between compiler and web workflows).}%
  \label{fig:ci-flow}
\end{figure}

Each workflow supports \texttt{workflow\_dispatch} for manual reruns and uses path filters so unrelated modules do not
rebuild on every commit.
Concurrency groups cancel superseded runs on the same branch.

\subsection{Deploy gate and responsibilities}\label{subsec:deploy-gate}

Production deploys on Railway are gated on successful GitHub CI for \texttt{main} (author-confirmed integration).
This chapter's deployment topology (Section~\ref{sec:deployment}) describes how passing pipelines promote Docker images
to Amsterdam-region services and how database migrations run before server rollout.

The distribution of tests matches the failure modes of a full-stack language tool: semantics and MIDI correctness in
C++; process and HTTP contracts on the server; parsing, rendering, and UX on the client; typographic and build integrity
on the paper.
Together with the architecture and feature sections above, this completes the engineering description of the
implemented system before the evaluation chapter measures compiler behavior quantitatively.
