\section{Comparative Interpretation}\label{sec:comparative-interpretation}

This section interprets the benchmark and implementation evidence from two angles.
First, it compares version~1.0 (implicit typing, \texttt{.dsl}) with version~2.0 (explicit typing, \texttt{.motivo}) on
identical workloads.
Second, it situates Motivo within the broader ecosystem of music programming tools.

\subsection{Version 1.0 versus Version 2.0}\label{detail-subsec:version-comparison}

The explicit-typing release changes the surface language and semantic pipeline without altering the lowering or MIDI
backend architecture.
Version~1.0 inferred variable types from initializers (\texttt{let}), resolved pattern overloads by arity alone, and ran
a second semantic pass (\texttt{validate\_pattern\_instantiation}) after the main traversal.
Version~2.0 requires explicit type annotations on variable declarations (\texttt{<type> IDENT = expr}), resolves
patterns by parameter type signatures, and
performs semantic analysis in a single traversal with assignability checks at declaration and assignment sites.

These design changes add compile-time guarantees: ill-typed programs fail before lowering, and pattern calls cannot
silently bind to the wrong overload when arities coincide but types differ.
The engineering question for this thesis is whether those guarantees impose measurable cost at scale.

Table~\ref{tab:benchmark-wall-clock} answers that question on the B1--B3 fixtures.
End-to-end wall-clock overhead of version~2.0 relative to version~1.0 is between $-0.08\%$ and $+1.74\%$---effectively
$0$--$2\%$ on an Apple M3 Pro Release build.
MIDI output sizes match byte-for-byte (Table~\ref{tab:benchmark-midi-size}), confirming that both compilers lower the
same workloads to equivalent artifacts.

The per-stage breakdown (Table~\ref{tab:benchmark-per-stage}) shows where the small difference arises.
Semantic time increases slightly in version~2.0 (for example B1: $+0.0009$~ms mean), which is expected because explicit
type annotations and signature-based pattern lookup add work per AST node.
That increase is dwarfed by lowering ($\sim$1.7--2.2~ms) and MIDI writing ($\sim$1.3--2.9~ms) on 20,000-event
programs.
Removing the second semantic pass in version~2.0 does not produce a dramatic speedup because both passes were already
sub-millisecond on these small ASTs; the win is primarily structural simplicity and earlier error reporting rather than
raw throughput.

In summary, the move from implicit to explicit typing trades a negligible wall-clock cost for stricter static guarantees
and a simpler semantic architecture, while preserving output equivalence on the benchmark corpus.

\subsection{Positioning within the Music Programming Landscape}\label{detail-subsec:positioning-dsl-ecosystem}

Representative music programming systems differ along three axes that also frame this evaluation: live performance
versus offline artifacts, composition separated from synthesis, and explicit temporal structure versus implicit
sequencing (Table~\ref{tab:related-work-comparison}).
Live-coding languages such as Sonic Pi, Tidal, Strudel, and HMusic prioritize interactive pattern manipulation and
stream events to external synthesis backends.
Real-time audio languages including ChucK and SuperCollider expose fine-grained timing and concurrent processes at the
cost of managing signal graphs and performance-time scheduling directly.
Notation and interchange formats such as ABC and MIDI encode musical information for typesetting or playback but lack
Motivo's combination of typed variables, parameterized patterns, and imperative control flow at the source level.

The benchmark evidence in this chapter does not reproduce those qualitative comparisons quantitatively.
It instead confirms the engineering side of Motivo's positioning: a zero-runtime compiler can expand compact structural
rules into thousands of MIDI events within the \texttt{MAX\_EVENTS} safety cap while incurring negligible overhead for
explicit typing in version~2.0.
Where Sonic Pi advances time through \texttt{sleep} and parallel \texttt{in\_thread} blocks during performance, and
Tidal and Strudel transform time-varying patterns at live-coding time, Motivo resolves voices and loops entirely at
compile time into a portable Standard MIDI File---the trade-off already summarized in
Table~\ref{tab:related-work-comparison} rather than re-derived here.
The B1--B3 fixtures stress that offline model through nested loops, chord flattening, and flat 20,000-event expansion
at the compiler cap.

Unlike raw MIDI scripting, Motivo keeps compositional
intent at the level of beats, patterns, and tracks; unlike ABC notation, the source is executable and procedurally
generative.
These distinctions motivated the language design; the measurements above show that pursuing them did not sacrifice
compile-time throughput on the tested workloads.

\subsection{Comparison with Raw MIDI Scripting}\label{detail-subsec:comparison-with-raw-midi-scripting}

For offline composition, many developers use general-purpose programming languages combined with MIDI libraries, such as
Python's \texttt{mido} or C++'s \texttt{RtMidi}.
While these tools offer absolute control over the output, they operate at an exceptionally low level of abstraction.

\topic{Event Placement and Musical Intent.}
When using a library like \texttt{mido}, the composer must manually construct \texttt{Note-On} and \texttt{Note-Off}
messages.
Crucially, because the MIDI file format relies on delta-times, the composer is forced to manually calculate the tick
difference between every sequential event across all tracks.

Motivo abstracts this entirely.
A composer writing \texttt{play (C4, E4, G4):2;} in Motivo communicates their \textit{musical intent}---playing a
C-Major triad for two beats.
Motivo's lowering engine and backend automatically handle the translation into absolute beats, the flattening into
individual notes, the sorting by absolute tick, and the final variable-length delta-time encoding.

\topic{Boilerplate and Expressiveness.}
A simple four-beat loop of a hi-hat playing eighth notes requires a significant amount of boilerplate in
Python: creating
a file, creating a track, appending tempo metadata, and writing a loop that interleaves \texttt{Note-On} and
\texttt{Note-Off} messages with explicit delta times.

Motivo achieves the same result with minimal syntax:
\begin{lstlisting}[language=Motivo,label={lst:midi-scripting-comparison}]
tempo 120;
track using drums {
    loop (8) { play hihat :0.5; }
}
\end{lstlisting}

This contrast in verbosity illustrates Motivo's intended role as a domain-specific tool.
By embedding musical semantics (like default durations, rest tracking, and automatic track/channel
assignment) directly into
the language design, the compiler reduces cognitive friction and allows the user to operate strictly within the musical
domain.
