\section{Methodology}\label{sec:evaluation-methodology}

The evaluation is driven by a versioned, reproducible benchmark suite in the \texttt{experiments/} module of the
repository.
A single command (\texttt{make run}) builds the required compilers, runs every workload, and regenerates the result
tables; the harness wraps each compiler invocation in Python's \texttt{time.perf\_counter}, performs 5 warmup runs, and
averages 50 measured runs.
Each measured command compiles one source file and writes a \texttt{.mid} file beside the input, so the end-to-end timer
includes process startup, parsing, semantic analysis, lowering, MIDI serialization, and file I/O.
All measurements reported in this chapter were taken on an Apple M3 Pro (arm64, macOS~26.5) with 18~GB of RAM, using
Release builds.

The benchmark programs fall into two groups.
\textit{Synthetic fixtures} push expansion and throughput toward the implementation's 20,000-event cap
(Section~\ref{sec:limits}) along several independent axes: a scaling sweep that varies only the number of generated
events, and structural programs that vary nesting depth, parallel tracks and voices, chord density, and per-note
expression load while holding the event count fixed.
\textit{Real compositions} are four complete pieces written for Motivo Studio, included to characterize behavior on
realistic, music-meaningful input rather than worst cases.
All programs are compiled by the current Motivo compiler, \texttt{motivoc}.
To isolate the cost of the version~2.0 type system, three fixtures that exist in both source dialects are additionally
compiled with the original version~1.0 build (\texttt{v1.0.0-base}, the \texttt{dslrc} CLI over \texttt{.dsl} sources)
alongside version~2.0 (\texttt{v2.0.0-explicit-types}, \texttt{motivoc} over \texttt{.motivo});
Section~\ref{detail-subsec:typing-cost} reports that comparison.

Compile time is decomposed into the four pipeline stages, parsing, semantic analysis, lowering, and MIDI serialization,
using in-process timers compiled in through the \texttt{MOTIVO\_ENABLE\_BENCHMARK} build option, which also emits the
lowered event count.
These timers report each stage's mean independently of subprocess overhead, so their sum is smaller than the end-to-end
wall-clock total, which additionally includes process startup and file I/O.
The instrumentation is gated at compile time, so production builds carry none of it.
The evaluation targets engineering properties (expansion ratio, compile time, scaling, and safety limits); musical
quality and user preference are out of scope.

\subsection{Theoretical Complexity}\label{detail-subsec:theoretical-complexity}

The compilation pipeline is composed of several sequential passes, each with a distinct computational complexity:

\topic{Frontend and Semantic Analysis.} The complexity of parsing and type-checking is proportional to the
number of lines of source code, $O(S)$, where $S$ is the size of the abstract syntax tree.
Because these passes do not unroll loops or instantiate patterns multiple times, they remain extremely fast regardless
of the composition's length.
Version~2.0 adds explicit type annotations and signature-based pattern resolution, but the traversal still visits each
AST node a constant number of times; the semantic pass was simplified from two traversals in version~1.0 to one.

\topic{Lowering Engine.} The unrolling process has a time complexity of $O(E)$, where $E$ is the total number
of generated \texttt{NoteEvent}s.
A simple \texttt{loop (1000)} statement will execute its body 1,000 times during this phase, meaning the lowering time
is proportional to the density of the final timeline, not the brevity of the source code.

\topic{MIDI Backend.} The backend must sort the absolute events by their tick values before computing
delta-times.
This stable sort introduces a time complexity of $O(E \log E)$.
Serializing the events into the binary file requires a final linear pass, $O(E)$.

Therefore, the overall time complexity of the compilation process is bounded by the sorting phase in the backend:
$O(E \log E)$.
