\section{Sensitivity to Program Shape}\label{sec:scaling-stress}

Section~\ref{sec:compile-time} established how compile time grows with the size of the generated timeline.
This section fixes the output near the cap and varies instead the \emph{shape} of the source program, to find which
language constructs make the same number of events more expensive to compile.
Each fixture emits 20{,}000 events (15{,}000 for the duration-expression case) but produces them differently: through
nesting depth, parallel tracks and voices, chord density, or per-note expression evaluation.
Table~\ref{tab:stress-suite} reports the full per-stage breakdown.

\begin{table}[htbp]
  \centering
  \footnotesize
  \setlength{\tabcolsep}{4.5pt}
  \renewcommand{\arraystretch}{1.25}
  \begin{tabular}{lrrrrrr}
    \toprule
    \textbf{Program shape} & \textbf{Events} & \textbf{Parse} & \textbf{Semantic} & \textbf{Lowering} &
    \textbf{MIDI write} & \textbf{End-to-end} \\
    \midrule
    Flat loop (baseline)         & 20,000 & 0.020 & 0.004 & 1.525 & 2.549 & $6.30 \pm 0.12$ \\
    Nested loops, 4 levels       & 20,000 & 0.022 & 0.004 & 2.107 & 2.566 & $6.93 \pm 0.12$ \\
    Nested loops, 5 levels       & 20,000 & 0.023 & 0.004 & 2.749 & 2.558 & $7.60 \pm 0.12$ \\
    20 parallel tracks           & 20,000 & 0.048 & 0.010 & 1.447 & 2.158 & $5.82 \pm 0.12$ \\
    10 parallel voices           & 20,000 & 0.038 & 0.006 & 1.576 & 2.445 & $6.29 \pm 0.13$ \\
    Four-note chords             & 20,000 & 0.024 & 0.004 & 1.252 & 2.511 & $6.06 \pm 0.31$ \\
    Eight-note chords            & 20,000 & 0.024 & 0.004 & 0.896 & 2.452 & $5.62 \pm 0.07$ \\
    Per-note duration arithmetic & 15,000 & 0.025 & 0.006 & 2.345 & 1.773 & $6.32 \pm 0.09$ \\
    Nested per-note ternaries    & 20,000 & 0.029 & 0.007 & 4.175 & 2.578 & $9.07 \pm 0.29$ \\
    \bottomrule
  \end{tabular}
  \caption{Per-stage mean compile time (ms) for fixtures of equal output size but differing source shape, current
    \texttt{motivoc}; the end-to-end column reports the mean $\pm$ standard deviation over 50 runs. Lowering varies with
  shape; MIDI write tracks the event count and is nearly constant.}\label{tab:stress-suite}
\end{table}

The breakdown separates two sources of cost.
MIDI write depends almost entirely on the event count: it stays near $2.5$~ms for every 20{,}000-event fixture
regardless of how the events were produced, because the backend sorts and serializes the same number of records in
every case.
The 15{,}000-event duration fixture and the eight-note-chord fixture write proportionally less, consistent with their
smaller or differently distributed event sets.
Lowering, in contrast, depends strongly on program shape, and it is where the differences between fixtures appear.

The largest factor is per-note expression evaluation.
The nested-ternary fixture re-evaluates a small arithmetic tree, three modulo operations and a nested conditional, for
every one of its 20{,}000 events, raising lowering to $4.18$~ms, about $2.7\times$ the flat baseline at the same event
count and making it the slowest workload measured anywhere in this chapter.
The per-note duration fixture shows the same mechanism on the duration path: at only 15{,}000 events its lowering still
reaches $2.35$~ms, roughly double the flat sweep's value at that size.
Nesting depth is the next largest factor.
Five levels of nested loops cost $2.75$~ms in lowering ($1.8\times$ the baseline) and four levels $2.11$~ms, because the
lowerer descends and restores more control-flow frames for each emitted event.
Chord density and parallelism, by contrast, stay close to the baseline: flattening four- or eight-note chords and
interleaving the onsets of twenty tracks or ten voices are all handled by the same per-event machinery and the backend
tick-sort without disproportionate cost.
The track- and voice-heavy fixtures show the only visible movement in the frontend, parsing rises to $0.05$~ms because
those sources are the longest (over a hundred lines), but the increase is still negligible against lowering and MIDI
write.

The conclusion is consistent across every shape: at a fixed output size, compile time is governed by lowering, and
within lowering by how much computation each emitted event demands, not by parsing or type checking.
Even the deliberately adversarial nested-ternary program compiles in about $9$~ms, comfortably within interactive
latency for the studio's compile-to-preview loop.
